---
title: "Project writeup"
author: "Tejas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F, echo = F)
options(scientific=T, digits = 3) 
```

```{r init libraries}
library(ezids)
library(tidyverse)
library(ggrepel)
```

```{r load data}
NYweath_Orig <- tibble(read.csv("data/NYC_weather_1869_2022.csv"))
NYweath <- NYweath_Orig %>%
  mutate(DATE = as.Date(DATE, format="%Y-%m-%d"),
                        MONTH = lubridate::month(DATE, label=T)) %>%
  filter(lubridate::year(DATE) >= 1900) %>%
  dplyr::select(TMAX, TMIN, DATE, PRCP, SNOW)
# view structure
str(NYweath)
# get summary
xkablesummary(NYweath)
```

## Temperature - stat tests of temp trends over time

We started by understanding the distributions of our temperature data to rule out any anomalies or unusual characteristics.  

```{r temperature distributions, results='asis'}
mean_tmax <- mean(NYweath$TMAX)
mean_tmin <- mean(NYweath$TMIN)

ggplot(NYweath) +
  geom_histogram(aes(x=TMAX, fill="Max Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmax, color="red", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmax + 9, y=2000, label=paste(round(mean_tmax, 2), "°F"), angle=0, size=4, color="darkred") +
  geom_histogram(aes(x=TMIN, fill="Min Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmin, color="blue", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmin - 9, y=2000, label=paste(round(mean_tmin, 2), "°F"), angle=0, size=4, color="darkblue") +
  labs(title="Distribution of Minimum and Maximum Daily Temperatures", x="Temperature (°F)", y="Count") +
  scale_y_continuous(position = "right") +
  scale_fill_manual(name="Column", values=c("Max Temp."="red","Min Temp."="blue"))
```

The distributions of both TMAX and TMIN do not appear to be normal at first glance, with TMAX being slightly skewed to the left and TMIN having two peaks. As expected, the distribution of TMAX includes temperature observations that are greater that those of TMIN, with a significant overlap. That overlap can be attributed to seasonality within the data. Because we have a large number of observations, we can assume normality for the sake of our statistical testing moving forward.  

```{r month average bar plot, results='asis'}
NYweath_Month <- NYweath %>%
  group_by(MONTH) %>%
  summarize("Avg Max Temp." = mean(TMAX, na.rm=T),
            "Avg Min Temp." = mean(TMIN, na.rm=T))

# side-by-side bar plot
NYweath_Month %>%
  dplyr::select(MONTH, "Avg Max Temp.", "Avg Min Temp.") %>%
  gather(key="Value", value="Temp.", "Avg Max Temp.", "Avg Min Temp.") %>%
  ggplot(aes(x=MONTH, y=Temp., fill=Value)) +
  geom_col(na.rm=TRUE, alpha=0.5, color="black", position="dodge") +
  labs(title="Average Daily Temperature By Month", x="Month", y="Average Temperature (°F)") +
  scale_fill_manual(values=c("red", "blue"))
```

The understand the effect of seasonality on the temperature data, we calculated the average daily temperatures by month. As we can see here, there is a trend of rising temperatures when we move from winter to summer months. This is to be expected since summer will see higher average daily maximum and minimum temperatures than winter. Seasonality is an aspect of our data that we might need to account for in our testing, as temperatures ranges clearly differ between summer and winter months.  

Now that we have insight into the distribution of our variables, it was time to start assessing whether the data confirmed any significant changes in daily temperatures over the time. Specifically, we want to test whether changes in average daily maximum temperatures coincide with the documented rise in global temperatures over the last century. A report on the global change in Earth's temperature and climate is available via the National Oceanic And Atmospheric Administration <https://www.ncdc.noaa.gov/sotc/global/202113>.  

We grouped the daily temperature observations into 3 distinct eras in American history. The eras are described as:  
  1. Industrial Era: 1900-1940  
  2. Cold War Era: 1940-1980  
  3. Modern Era: 1980-Present  
  
This distinction was important because we could now compare temperature observations across three independent groups, and each era contains roughly 15,000 daily observations giving us statistically powerful sample sizes to work off of.   
 
```{r era distributions}
industrial_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1900, lubridate::year(DATE) < 1940)

coldwar_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1940, lubridate::year(DATE) < 1980)

modern_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1980)
```


```{r era box plots, results='asis'}
# modify data set by adding ERA column
industrial_era <- industrial_era %>%
  mutate(Era = 1)
coldwar_era <- coldwar_era %>%
  mutate(Era = 2)
modern_era <- modern_era %>%
  mutate(Era = 3)
# combine data frames
all_eras <- rbind(industrial_era, coldwar_era, modern_era)
# convert ERA column to factor/categorical
all_eras$Era <- factor(all_eras$Era, labels = c("Indst.", "Cold War", "Modern"))

# box plot
means <- aggregate(TMAX ~ Era, all_eras, mean)

all_eras %>%
  ggplot(aes(x=Era, y=TMAX, fill=Era)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  stat_summary_bin(fun="mean", geom="point", shape=20, size=5, show.legend=FALSE) +
  geom_text(data = means, aes(label=round(TMAX, 2) , y=TMAX - 5)) +
  labs(title="Distribution of Daily Maximum Temperature By Era", x="Era", y="Temperature (°F)") +
  scale_fill_brewer(palette="Dark2")
```

The distribution of each era looks similar, however the calculated average daily maximum temperatures suggests there's a slight difference across the time periods. To verify the statistical significance of these differences, we ran an ANOVA test comparing the three eras.  

```{r anova and tukey}
# run ANOVA
anova_res <- aov(TMAX ~ Era, data = all_eras)
xkabledply(anova_res)
```

With a p-value much lower than our threshold of 0.05, we have verified that the statistical differences between the average daily maximum temperature of the 3 eras is significant.  

That's good news, but just how difference are they from each other? For that answer, we can turn to Tukey's HSD (honestly significant difference) test. This test helps us compare the differences between all possible pairs of our groupings and evaluate their significance.  

```{r tukey test, results='asis'}
tukeyAoV <- TukeyHSD(anova_res, conf.level=0.95)
plot(tukeyAoV, col="red", cex.axis=0.75)
```
The plot displays the estimated difference between the means of each dual-era comparison along with a 95% confidence interval for that difference. The adjusted p-value for each 2-way comparison was practically 0 which indicates that the differences in the average daily maximum temperature between the eras are significant.  

Through this analysis, we are able to safely reject the hypothesis that the average daily maximum temperature remains the same across each era. There is significant evidence to suggest that our TMAX variable is increasing over time. We can utilize linear regressions to dive further into this relationship.  


## Precipitation trends -Tejas' outlier analysis + snowfall at Christmas

As average temperatures on Earth rise, more evaporation occurs which increases overall precipitation. To gauge the effect of this on the local level, we sought to analyze precipitation outliers in our dataset and determine if a significant change has occurred in rainfall over time.  

Outliers are defined as values that lie significantly far from other points in the data. Because there were a lot of days with 0 inches of precipitation that skewed our data, we decided to remove them before calculating our outliers.  

```{r prcp outliers}
# extract PRCP outlier rows into df variable
# remove zeros
# group variable by decade, getting sum for each decade
prcp_no_zeros <- NYweath %>%
  filter(PRCP > 0.0 & !is.na(PRCP))
prcp_boxplot <- boxplot(prcp_no_zeros$PRCP, ylab = "Temperature (°F)")

prcp_outliers <- boxplot.stats(prcp_no_zeros$PRCP)$out
prcp_outliers.index <- which(prcp_no_zeros$PRCP %in% c(prcp_outliers))
prcp_outliers_df <- prcp_no_zeros[prcp_outliers.index, ]

prcp_outliers_df.yearly <- prcp_outliers_df %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarize(prcp_total = sum(PRCP, na.rm=T), count = n())

# show top 5 years
top_5_years.prcp_total <- prcp_outliers_df.yearly %>%
  slice_max(order_by = prcp_total, n = 5)

# group into decades and create bar plot
prcp_outliers_df.decades <- prcp_outliers_df %>%
  filter(lubridate::year(DATE) < 2020) %>%
  mutate(decade = paste(floor(lubridate::year(DATE) / 10) * 10, "s", sep="")) %>%
  group_by(decade) %>%
  summarise(prcp_total = sum(PRCP, na.rm=T), count = n())
```
```{r prcp outlier plots, results='asis'}
prcp_outliers_df.yearly %>%
  ggplot(aes(x=YEAR, y=prcp_total)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="darkgreen") +
    geom_point(na.rm = TRUE, fill="#69b3a2", shape = 21) +
    geom_text_repel(aes(label=YEAR), data=top_5_years.prcp_total) +
    labs(title="Total Precipitation By Year", x="Year", y="Total Yearly Amount (inches)")
```
Looking at the yearly data, we identified a discernible trend of increased variation within the amount of total rainfall within a year starting after 1970. The top 5 years based on total precipitation all reside within the previous half century of data.  

```{r prcp decades, results='asis'}
prcp_outliers_df.decades %>%
  ggplot(aes(group=1)) +
  geom_col(aes(x=decade, y=prcp_total), na.rm=TRUE, alpha=0.8, color="black", fill="darkgreen", position="identity") +
  geom_line(aes(x=decade, y=count), stat="identity", color="#69b3a2", size=1) +
  labs(title="Precipitation Outlier Days And Total Precipitation By Decade", x="Decade", y="Total Amount (inches)") +
  scale_y_continuous(sec.axis=sec_axis(~., name="Days with Outliers"))
```

Zooming out, we can see a relationship forming between the amount of days with precipitation outlier values and the total precipitation amount occurring over the course of a decade. We see an upwards trend in both values starting in the 1970s, confirming the visual trends analyzed in the yearly totals plot above.  

This coincides with the patterns of increasing average daily temperature we saw in our prior analysis. Although not part of this report, we could further investigate the correlation between temperature and precipitation to determine if the data signals a statistically significant pattern.  

Another indicator of warming trends, especially for a location such as New York City that experiences all 4 seasons regularly, is the amount of snowfall occurring in a given year. In this dataset, there were an unusual amount of missing snowfall observations which made performing this analysis difficult. What we did have consistent values on was snowfall over the Christmas holiday (Christmas Eve and Christmas Day).  

```{r snowfall, results='asis'}
nyc_weather.christmas_snow <- NYweath %>%
  filter(lubridate::month(DATE) == 12, (lubridate::day(DATE) == 24 | lubridate::day(DATE) == 25)) %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarise(snowfall = sum(SNOW, na.rm = T))

# how many years has there been snow?
with_snow <- nyc_weather.christmas_snow %>%
  filter(snowfall > 0.0)

# yearly plot for years it snowed
with_snow %>%
  ggplot(aes(x=YEAR, y=snowfall)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="steelblue", size=1.25) +
    geom_point(na.rm = TRUE, fill="#00edfa", shape = 21) +
    geom_label_repel(aes(label=YEAR), data=with_snow) +
    labs(title="Years With Snowfall Over Christmas Holiday", x="Year", y="Snowfall (inches)")
```

Since 1900, it has snowed on 24 out of 121 Christmas holidays (~20%). Analyzing the plot above, we can see a trend of reducing snowfall events over the Christmas holiday beginning around 1980. Between 1980 and 2021, there have only been 4 Christmas holiday where snowfall occurred (~10%). If we are able to acquire the necessary data, we can further test the significance of our observations to determine if a statistical correlation between snowfall and temperature exists.