---
title: "Project writeup"
author: "Emily"
date: "2022-10-29"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

output: html_document: code_folding: hide number_sections: false toc:
yes toc_depth: 3 toc_float: yes pdf_document: toc: yes toc_depth: '3'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

#loading packages 
library(ggplot2)
library(ezids)
library(dplyr)
```

## Emily's starter Code

```{r, results="asis"}
#loading data 

NYweath <- data.frame(read.csv("data/NYC_weather_1869_2022.csv"))
#xkablesummary(NYweath)
JFKweath <- data.frame(read.csv("data/JFK_weather_1949-2022.csv"))
#xkablesummary(JFKweath)

```

```{r, results="asis"}
#loading packages and data 

NYweath <- data.frame(read.csv("data/NYC_weather_1869_2022.csv"))
#xkablesummary(NYweath)
#str(NYweath)                      
#cablehead(NYweath, n=5)
#tail(NYweath, n=2)
#nrow(NYweath)

```

```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
NYweath$DATE <- as.Date(NYweath$DATE)
NYweath$day <- format(NYweath$DATE, format="%d")
NYweath$month <- format(NYweath$DATE, format="%m")
NYweath$year <- format(NYweath$DATE, format="%Y")

#converting temperature observations to numerics
NYweath$TMAX <- as.numeric(NYweath$TMAX)
NYweath$TMIN <- as.numeric(NYweath$TMIN)
NYweath$TAVG <- as.numeric(NYweath$TAVG)
NYweath$year <- as.numeric(NYweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data that are only date- and temperature-related
NYweath_sub <- subset(NYweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 

#creating a subset for 1900 on
NYweath_00 <- subset (NYweath_sub, year > 1899) # subsetting to 20th century and later
xkabledplyhead(NYweath_00)

#creating a subset for 1948 on for comparison with JFK data
NYweath_48 <- subset (NYweath_sub, year > 1947)

#now, same for JFK data  

#converting to R date format and adding columns for day, month, and year  
JFKweath$DATE <- as.Date(JFKweath$DATE)
JFKweath$day <- format(JFKweath$DATE, format="%d")
JFKweath$month <- format(JFKweath$DATE, format="%m")
JFKweath$year <- format(JFKweath$DATE, format="%Y")

#converting temperature observations to numerics
JFKweath$TMAX <- as.numeric(JFKweath$TMAX)
JFKweath$TMIN <- as.numeric(JFKweath$TMIN)
JFKweath$TAVG <- as.numeric(JFKweath$TAVG)
JFKweath$year <- as.numeric(JFKweath$year)

#Making month a factor
JFKweath$month <- as.factor(JFKweath$month)

#creating a subset of the weather data that are only date- and temperature-related
JFKweath_sub <- subset(JFKweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(JFKweath_sub)
```

##Introduction

As scientists, our team is curious about how humans can affect the
natural environment. We wondered if weather conditions in a highly
populated region might correlate with notable human activities on a
global or local scale.

There is a variety of literature available on global climate change,
suggesting that the Earth's surface temperature is increasing, on
average, over time in a way that threatens many lifeforms on our planet.
This warming is primarily due to human activity: the combustion of
fossil fuels resulting in the emission of carbon dioxide into the
atmosphere. We decided to consider local weather trends in a specific
urban area to see what we might learn.

After some search, we discovered that historical daily weather records
from locations around the United States are made publicly availably by
the National Oceanic and Atmospheric Administrative via their website:
<https://www.ncdc.noaa.gov/cdo-web>. We found that a data station in the
middle of Central Park in New York City has made more than 56,000 daily
weather observations dating back to 1869. The variables observed
included daily maximum temperature in degrees Fahrenheit (TMAX), daily
minimum temperature in degrees Fahrenheit (TMIN), and daily precipitation (PRCP) and snowfall (SNOW) in inches. We decided to analyze
these data to see what trends we might uncover. Our data span the days
from February 28, 1869, to September 26, 2022.

Temperature data from Central Park have been studied in the past, and
warming trends were observed; we wanted to see what we might discover on
our own, from analysis of the original data.

## Our Research Questions

Given that Central Park is an oasis in the middle of a city environment,
we wondered whether any weather trends correlate with major human
activities-- both globally and locally. To explore these possibilities,
we formulated the following preliminary questions:

1.  Are there statistically measurable changes in weather patterns
    (e.g., temperature and precipitation levels) over time in New York
    City?\
2.  Do temperature trends in New York City align with the documented
    rise in average global temperatures over the last century?\
3.  Are there any notable (and statistically significant) changes in
    weather patterns after/during the pandemic lockdown (perhaps due to
    changes in commuting patterns)?

To dig in, we began with temperature.

## Temperature - stat tests of temp trends over time

## Chris's linear model - blurb

## Emily's linear model Linear Models of Temperature over Time

We then created a linear model of TMAX vs. year to understand the temperature trends since 1900. The fit parameters were statistically significant, and suggested that both the maximum and minimum daily temperatures in New York's Central Park have increased on average over time at a rate of approximately 0.026 degrees per year. While the p-value for this parameter is < 2e-16 (well below the threshold of alpha = 0.05), this overall fit is poor, with an adjusted r-squared value of 0.00245.  


```{r}
# Linear regression of TMAX \~ year from 1900 on.
maxTfit00_0 <- lm(formula = TMAX ~ year, data = NYweath_00 )
summary(maxTfit00_0)  
```

The poor fit is likely due to the wide range of daily temperatures that occur in a given year as a result of seasonal variation. The following plot of daily maximum temperatures shows the wide variance of the data around the linear model. 

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("01" = "purple4",
                                  "02" = "purple",
                                  "03" = "dodgerblue",
                                  "04" = "cyan4",
                                  "05" = "yellow",
                                  "06" = "darkgoldenrod1",
                                  "07" = "red",
                                  "08" = "orange", 
                                  "09" = "yellow",
                                  "10" = "green",
                                  "11" = "turquoise",
                                  "12" = "steelblue")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)  
```


## Linear regression of TMAX \~ year + month from 1900 on.

```{r}
# Linear regression of TMAX \~ year from 1900 on.
maxTfit00_0 <- lm(formula = TMIN ~ year, data = NYweath_00 )
summary(maxTfit00_0)  
```

In order to improve the fit and model temperature trends more completely, we decided to account for seasonal variation by also including month as a categorical regressor. The resulting fit has an r-squared value of 0.775 and a slope of 0.025 degrees Fahreinheit per year, with all fit parameters' p-values well below 0.05. The different intercepts for the each level of the categorical variable (the twelve months of the year) indicate that January is the coldest and July the hottest month in Central Park, with an average difference in maximum daily temperature of approximately 46 degrees Fahrenheit in any given year over this window.  


```{r}
maxTfit00_1 <- lm(formula = TMAX ~ year + month, data = NYweath_00 )
summary(maxTfit00_1)  

```

These two extremes and their linear models are plotted in the following figure; it is clear that the multiple regression is a much better model of temperature trends, consistent with the higher r-squared value.  


```{r}
ggplot(NYweath_00, aes(x = year, y = TMAX, color = month)) +
    geom_point() +
    scale_color_manual(values = c("01" = "purple4",
                                   "07" = "red"), na.value = NA) +
    geom_abline(aes(intercept = -11.05508, slope = 0.02539), col = "black", size = 1) + 
    geom_abline(aes(intercept = 34.98295, slope = 0.02539), col = "black", size = 1) +
  
    labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park")  
#    stat_smooth(method = "lm",
#            formula = y ~ x,
#            geom = "smooth",
#            color = "black",
#            show.legend = F)
```

To create an even better model, we'd need to use sinusoidal functions that capture the cyclical variation of weather with season.


## Central Park v. Global average - Chris

## JFK data, comparable? - Emily

We wondered whether these trends were true for other locations in the New York City area. To assess this, we found data from another NOAA station at JFK International Airport. Because these data date only as far back as the Airport (which was built in 1948), we focused on 1948 on, computing linear models for both regions for this time window.  

We were suprised to notice that the slope of the Central Park model for 1948 on was lower than that including observations from 1900 on; only 0.014 degrees per year compared to 0.025. This suggests that average Central Park warming was greater in the first half of the 20th century than in the second half-- which is not what we would intuit based on the understanding that the global rate of warming is increasing.  

We also found a higher warming rate at the JFK airport sight, of approximately 0.033 degrees Fahrenheit per year.  

```{r, echo = FALSE}

#Linear models from 1948 on

maxTfit48_1 <- lm(formula = TMAX ~ year + month, data = NYweath_48 )
summary(maxTfit48_1)

# Multiple linear regression for JFK (from 1948 on  
jmaxTfit1 <- lm(formula = TMAX ~ year + month, data = JFKweath_sub )
summary(jmaxTfit1)  

```

To see whether the different warming rates in Central Park post-1900 and post-1948, and at JFK airport are real, we examined the 95 percent confidence intervals associated with each of the three slopes.  

```{r}
warming_compare <- data.frame(col1 = c("Central Park", "Central Park", "JFK"), col2 = c("1900", "1948", "1948"), col3 = c("x", "y", "z") )

```


How did thinking change? reframing of questioning?\
Additional data to test our hypothesis of causality

## Precipitation trends -Tejas' outlier analysis + snowfall at Christmas

## Linear models of precipitation trends over time - Chris

## Correlation between temp / precip -- Emily ... think about subsequent analyses

## Lockdown - Chris

## Conclusion- terse summary of findings, restate new questions from all sections, ideas for future analyses if you give us more money.

If we have time, summarize what we learned from each variable, and new
questions?
