---
title: "Project writeup"
author: "Tejas"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F, echo = F)
options(scientific=T, digits = 3) 
```

```{r init libraries}
library(ezids)
library(tidyverse)
library(ggrepel)
```

```{r load data}
NYweath_Orig <- tibble(read.csv("data/NYC_weather_1869_2022.csv"))
NYweath <- NYweath_Orig %>%
  mutate(DATE = as.Date(DATE, format="%Y-%m-%d"),
                        MONTH = lubridate::month(DATE, label=T)) %>%
  filter(lubridate::year(DATE) >= 1900) %>%
  dplyr::select(TMAX, TMIN, DATE, PRCP, SNOW)
# view structure
str(NYweath)
# get summary
xkablesummary(NYweath)
```

## Temperature - stat tests of temp trends over time

We started by understanding the distributions of our temperature data to rule out any anomalies or peculiar characteristics. Another aspect of our data that we could potentially account for is seasonality, as temperatures ranges differ between summer and winter months.

```{r temperature distributions, results='asis'}
mean_tmax <- mean(NYweath$TMAX)
mean_tmin <- mean(NYweath$TMIN)

ggplot(NYweath) +
  geom_histogram(aes(x=TMAX, fill="Max Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmax, color="red", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmax + 9, y=2000, label=paste(round(mean_tmax, 2), "°F"), angle=0, size=4, color="darkred") +
  geom_histogram(aes(x=TMIN, fill="Min Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmin, color="blue", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmin - 9, y=2000, label=paste(round(mean_tmin, 2), "°F"), angle=0, size=4, color="darkblue") +
  labs(title="Distribution of Minimum and Maximum Daily Temperatures", x="Temperature (°F)", y="Count") +
  scale_y_continuous(position = "right") +
  scale_fill_manual(name="Column", values=c("Max Temp."="red","Min Temp."="blue"))
```

The distributions of both TMAX and TMIN do not appear to be normal at first glance, with TMAX being slightly skewed to the left and TMIN having two peaks. The average daily maximum temperature is `r paste(round(mean_tmax, 2), "°F")`, and the average daily minimum temperature is `r paste(round(mean_tmin, 2), "°F")`. As expected, the distribution of TMAX includes temperature observations that are greater that those of TMIN, with a significant overlap. That overlap can be attributed to seasonality within the data. Because we have a large number of observations, we can assume normality for the sake of our statistical testing moving forward.  

```{r month average bar plot, results='asis'}
NYweath_Month <- NYweath %>%
  group_by(MONTH) %>%
  summarize("Avg Max Temp." = mean(TMAX, na.rm=T),
            "Avg Min Temp." = mean(TMIN, na.rm=T))

# side-by-side bar plot
NYweath_Month %>%
  dplyr::select(MONTH, "Avg Max Temp.", "Avg Min Temp.") %>%
  gather(key="Value", value="Temp.", "Avg Max Temp.", "Avg Min Temp.") %>%
  ggplot(aes(x=MONTH, y=Temp., fill=Value)) +
  geom_col(na.rm=TRUE, alpha=0.5, color="black", position="dodge") +
  labs(title="Average Daily Temperature By Month", x="Month", y="Average Temperature (°F)") +
  scale_fill_manual(values=c("red", "blue"))
```

The understand the effect of seasonality on the temperature data, we calculated the average daily temperatures by month. As we can see, there is a trend of rising temperatures when we move from the colder, winter months to the warmer, summer months. This is to be expected since summer will have higher average daily maximum and minimum temperatures than winter.  

Now that we have insight into the distribution of our variables, it was time to start assessing whether the data confirmed any significant changes in daily temperatures over the timeline of our dataset. Specifically, we want to test whether changes in average daily maximum temperatures coincide with the documented rise in global temperatures over the last century. A report on the global change in Earth's temperature and climate is available via the National Oceanic And Atmospheric Administration <https://www.ncdc.noaa.gov/sotc/global/202113>.  

#### Results of Era analysis

In an effort to categorize our `DATE` variable, we grouped the daily temperature observations into 3 distinct eras in American history. The eras are described as:  
  1. Industrial Era: 1900-1940  
  2. Cold War Era: 1940-1980  
  3. Modern Era: 1980-Present (2022)  
  
This distinction was important because we could now compare temperature observations across three different and unique samples. The subset our data into the different eras, we extracted the year of the observation from the `DATE` column and filtered based on the era time range.  
 
```{r era distributions}
industrial_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1900, lubridate::year(DATE) < 1940)

coldwar_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1940, lubridate::year(DATE) < 1980)

modern_era <- NYweath %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1980)
```

Each era contains roughly 15,000 daily observations which is an important factor in the analysis we will perform next.  

The distribution of each era is plotted below:  
```{r era box plots, results='asis'}
# modify data set by adding ERA column
industrial_era <- industrial_era %>%
  mutate(Era = 1)
coldwar_era <- coldwar_era %>%
  mutate(Era = 2)
modern_era <- modern_era %>%
  mutate(Era = 3)
# combine data frames
all_eras <- rbind(industrial_era, coldwar_era, modern_era)
# convert ERA column to factor/categorical
all_eras$Era <- factor(all_eras$Era, labels = c("Indst.", "Cold War", "Modern"))

# box plot
means <- aggregate(TMAX ~ Era, all_eras, mean)

all_eras %>%
  ggplot(aes(x=Era, y=TMAX, fill=Era)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  stat_summary_bin(fun="mean", geom="point", shape=20, size=5, show.legend=FALSE) +
  geom_text(data = means, aes(label=round(TMAX, 2) , y=TMAX - 5)) +
  labs(title="Distribution of Daily Maximum Temperature By Era", x="Era", y="Temperature (°F)") +
  scale_fill_brewer(palette="Dark2")
```

The calculated average daily maximum temperatures for each era are as follows:  
  1. Industrial Era: `60.91` $^{\circ}F$  
  2. Cold War Era: `62.25` $^{\circ}F$  
  3. Modern Era: `63.06` $^{\circ}F$  
  
#### Statistical Test
  
These results indicate a change in average daily maximum temperatures between the eras. To test the statistical significance of these difference we ran an ANOVA test comparing the three eras.

```{r anova and tukey, results='asis'}
# run ANOVA
anova_res <- aov(TMAX ~ Era, data = all_eras)
xkabledply(anova_res)
```

With a p-value of `r pf(51, 2, 44826, lower.tail=F)`, we have verified that the statistical differences between the average daily maximum temperature of the 3 eras is significant. That's good news, but just how difference are they from each other? For that answer, we can turn to Tukey's HSD (honestly significant difference) test. This test helps us compare the differences between all possible pairs of eras and evaluate their significance.  

```{r tukey test, results='asis'}
tukeyAoV <- TukeyHSD(anova_res, conf.level=0.95)
plot(tukeyAoV, col="red", cex.axis=0.75)
```
The test results confirm that the difference in average daily maximum temperature between each pair of eras is significantly difference. The plot helps highlights these differences.  

Through our analysis of the observations between the three eras, we are able to safely reject the hypothesis that the average daily maximum temperature remains the same. There is significant evidence to suggest that our TMAX variable is increasing over time.  




## Precipitation trends -Tejas' outlier analysis + snowfall at Christmas

