---
title: "Project 1 EDA: Central Park Weather Analysis"
author: "Christopher Washer, Emily Grumbling, Tejas Rawal"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# tidyverse includes ggplot2
library(tidyverse)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

# NYC Weather Data: Temperature analysis

```{r load data}
nyc_weather <- tibble(read.csv("../data/NYC_weather_1869_2022.csv")) %>%
  dplyr::select(TMAX, TMIN, DATE, PRCP, SNOW)
# view structure
str(nyc_weather)
# get summary
xkablesummary(nyc_weather)
```

Convert DATE column from `char` to `Date` type
```{r}
nyc_weather <- nyc_weather %>%
  mutate(DATE = as.Date(DATE, format="%Y-%m-%d"))

# review dataset structure
xkablesummary(nyc_weather)
```

Add a MONTH column to the dataset. The month is extracted from the DATE.  
Subset data for DATE > 1900
```{r}
nyc_weather <- nyc_weather %>%
  filter(lubridate::year(DATE) >= 1900) %>%
  mutate(MONTH = lubridate::month(DATE, label=T))

str(nyc_weather)
# review dataset structure
xkablesummary(nyc_weather)
```

### Temperature Plots
```{r}

mean_tmax <- mean(nyc_weather$TMAX)
mean_tmin <- mean(nyc_weather$TMIN)

ggplot(nyc_weather) +
  geom_histogram(aes(x=TMAX, fill="Max Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmax, color="red", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmax + 9, y=2000, label=paste(round(mean_tmax, 2), "°F" ), angle=0, size=4, color="darkred") +
  geom_histogram(aes(x=TMIN, fill="Min Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmin, color="blue", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmin - 9, y=2000, label=paste(round(mean_tmin, 2), "°F" ), angle=0, size=4, color="darkblue") +
  labs(title="Distribution of Minimum and Maximum Daily Temperatures", x="Temperature (°F)", y="Count") +
  scale_y_continuous(position = "right") +
  scale_fill_manual(name="Column", values=c("Max Temp."="red","Min Temp."="blue"))

nyc_weather %>%
  ggplot(aes(y=TMAX)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black", fill="red") +
  labs(title="TMAX Boxplot", y="Temperature (°F)")

nyc_weather %>%
  ggplot(aes(y=TMIN)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black", fill="steelblue") +
  labs(title="TMIN Boxplot", x="", y="Temperature (°F)")


qqnorm(nyc_weather$TMAX,
       main="QQ Plot TMAX",
       xlab="Theoretical Quantile",
       ylab="Temperature  (°F)",
       col="red")
qqline(nyc_weather$TMAX, lwd=1, lty=2)

qqnorm(nyc_weather$TMIN,
       main="QQ Plot TMIN",
       xlab="Theoretical Quantile",
       ylab="Temperature  (°F)",
       col="blue")
qqline(nyc_weather$TMIN, lwd=1, lty=2)
```

### Average daily temperature max and min by month, all time
```{r average daily temperature analysis}
# group data by month
# and calculate average
nyc_weather_by_month <- nyc_weather %>%
  group_by(MONTH) %>%
  summarize("Avg Max Temp." = mean(TMAX, na.rm=T),
            "Avg Min Temp." = mean(TMIN, na.rm=T),
            avg_prcp = mean(PRCP, na.rm=T),
            avg_snow = mean(SNOW, na.rm=T))

# side-by-side bar plot
nyc_weather_by_month %>%
  dplyr::select(MONTH, "Avg Max Temp.", "Avg Min Temp.") %>%
  gather(key="Value", value="Temp.", "Avg Max Temp.", "Avg Min Temp.") %>%
  ggplot(aes(x=MONTH, y=Temp., fill=Value)) +
  geom_col(na.rm=TRUE, alpha=0.5, color="black", position="dodge") +
  labs(title="Average Daily Temperature By Month", x="Month", y="Average Temperature (°F)") +
  scale_fill_manual(values=c("blue", "red"))
```


### Average percipitation and snow fall by month, all time
```{r}
nyc_weather_by_month %>%
  dplyr::select(MONTH, avg_prcp, avg_snow) %>%
  gather(key = "data_point", value="value", avg_prcp, avg_snow) %>%
  ggplot(aes(x=MONTH, y=value, fill=data_point)) +
    geom_col(na.rm=TRUE, alpha=0.7, color="black", position="dodge") +
    labs(x="Month", y="Amount (inches)")
```

## Categorical Analysis

For this analysis, I sought to subset the data into 30-50 year periods as such:  
* 1900-1940: Industrial Revolution  
* 1940-1980: Cold War Era  
* 1980-present: Modern Era

### 1900-1940
#### Temperature and Precipitation Boxplots
```{r}
# filter weather data for years 1900-1940
industrial_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1900, lubridate::year(DATE) < 1940)

str(industrial_era)

industrial_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1900-1940", x="Temperature (°F)", y="Count")

industrial_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1900-1940", x="Data Point", y="Temperature (°F)")

industrial_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1900-1940", x="Data Point", y="Amount (inches)")

```

### 1940-1980
#### Temperature and Precipitation Boxplots
```{r era-based temperature analysis}
coldwar_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1940, lubridate::year(DATE) < 1980)

str(coldwar_era)

coldwar_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1940-1980", x="Temperature (°F)", y="Count")

coldwar_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1940-1980", x="Data Point", y="Temperature (°F)")

coldwar_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1940-1980", x="Data Point", y="Amount (inches)")

```

### 1980 - present
#### Temperature and Precipitation Boxplots
```{r}
modern_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1980)

str(modern_era)

modern_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1980-Present", x="Temperature (°F)", y="Count")

modern_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1980-Present", x="Data Point", y="Temperature (°F)")

modern_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1980-Present", x="Data Point", y="Amount (inches)")
```

### ANOVA test

#### Mean TMAX by era
```{r}
industrial_avg_tmax <- mean(industrial_era$TMAX)
coldwar_avg_tmax <- mean(coldwar_era$TMAX)
modern_avg_tmax <- mean(modern_era$TMAX)

industrial_avg_tmax
coldwar_avg_tmax
modern_avg_tmax
```

We can see above that there is an increase in average daily temperature max over the three eras.
$H_0:$ The average `TMAX` across the three eras is equal.  
$H_1:$ The average `TMAX` across the three eras is different.  
$\alpha$: `0.05`

```{r}
# modify data set by adding ERA column
industrial_era <- industrial_era %>%
  mutate(Era = 1)
coldwar_era <- coldwar_era %>%
  mutate(Era = 2)
modern_era <- modern_era %>%
  mutate(Era = 3)
# combine data frames
all_eras <- rbind(industrial_era, coldwar_era, modern_era)
# convert ERA column to factor/categorical
all_eras$Era <- factor(all_eras$Era, labels = c("Indst.", "Cold War", "Modern"))

# box plot
means <- aggregate(TMAX ~ Era, all_eras, mean)

all_eras %>%
  ggplot(aes(x=Era, y=TMAX, fill=Era)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  stat_summary_bin(fun="mean", geom="point", shape=20, size=5, show.legend=FALSE) +
  geom_text(data = means, aes(label=round(TMAX, 2) , y=TMAX - 5)) +
  labs(title="Distribution of Daily Maximum Temperature By Era", x="Era", y="Temperature (°F)") +
  scale_fill_brewer(palette="Dark2")
```

```{r}
# run ANOVA
anova_res <- aov(TMAX ~ Era, data = all_eras)
summary(anova_res)
xkabledply(anova_res)
```

With a p-value of `r pf(51, 2, 44826, lower.tail=F)`, we can safely assume that the average daily maximum temperature between the 3 eras is different.

#### Tukey HSD

This tests will analyze the difference between means of the 3 eras:  

```{r}
tukeyAoV <- TukeyHSD(anova_res, conf.level=0.95)
tukeyAoV
plot(tukeyAoV, col="red", cex.axis=0.75)
```

### Outlier Check on Precipitation
```{r}
library(ggrepel)
# extract PRCP outlier rows into df variable
# remove zeros
# group variable by decade, getting sum for each decade
prcp_no_zeros <- nyc_weather %>%
  filter(PRCP > 0.0 & !is.na(PRCP))
prcp_boxplot <- boxplot(prcp_no_zeros$PRCP, ylab = "Temperature (°F)")

prcp_outliers <- boxplot.stats(prcp_no_zeros$PRCP)$out
prcp_outliers.index <- which(prcp_no_zeros$PRCP %in% c(prcp_outliers))
prcp_outliers_df <- prcp_no_zeros[prcp_outliers.index, ]

prcp_outliers_df.yearly <- prcp_outliers_df %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarize(prcp_total = sum(PRCP, na.rm=T), count = n())

# show top 5 years
top_5_years.prcp_total <- prcp_outliers_df.yearly %>%
  slice_max(order_by = prcp_total, n = 5)

prcp_outliers_df.yearly %>%
  ggplot(aes(x=YEAR, y=prcp_total)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="darkgreen") +
    geom_point(na.rm = TRUE, fill="#69b3a2", shape = 21) +
    geom_text_repel(aes(label=YEAR), data=top_5_years.prcp_total) +
    labs(x="Year", y="Total Yearly Amount (inches)") +
    scale_fill_brewer(palette="Spectral")

top_5_years.count <- prcp_outliers_df.yearly %>%
  slice_max(order_by = count, n = 5)

prcp_outliers_df.yearly %>%
  ggplot(aes(x=YEAR, y=count)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="darkgreen") +
    geom_point(na.rm = TRUE, fill="#69b3a2", shape = 21) +
    geom_text_repel(aes(label=YEAR), data=top_5_years.count) +
    labs(title="Precipitation Outlier Days By Year", x="Year", y="Days with Outliers")

# group into decades and create bar plot
prcp_outliers_df.decades <- prcp_outliers_df %>%
  mutate(decade = paste(floor(lubridate::year(DATE) / 10) * 10, "s", sep="")) %>%
  group_by(decade) %>%
  summarise(prcp_total = sum(PRCP, na.rm=T), count = n())

prcp_outliers_df.decades %>%
  ggplot(aes(group=1)) +
  geom_col(aes(x=decade, y=prcp_total), na.rm=TRUE, alpha=0.8, color="black", fill="darkgreen", position="identity") +
  geom_line(aes(x=decade, y=count), stat="identity", color="#69b3a2", size=1) +
  labs(title="Precipitation Outlier Days And Total Precipitation By Decade", x="Decade", y="Total Amount (inches)") +
  scale_y_continuous(sec.axis=sec_axis(~., name="Days with Outliers"))

prcp_outliers_df.decades %>%
  ggplot(aes(x=decade, y=count)) +
    geom_col(na.rm=TRUE, alpha=0.7, color="black", fill="#56B4E9", position="identity") +
    labs(x="Decade", y="Days with Outliers")
```

### Christmas Snow
```{r}
# sum snowfall from 24-25th Dec
nyc_weather.christmas_snow <- nyc_weather %>%
  filter(lubridate::month(DATE) == 12, (lubridate::day(DATE) == 24 | lubridate::day(DATE) == 25)) %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarise(snowfall = sum(SNOW, na.rm = T))

# how many years has there been snow?
with_snow <- nyc_weather.christmas_snow %>%
  filter(snowfall > 0.0)

# highest snowfall year?
largest_snowfall <- nyc_weather.christmas_snow[which.max(nyc_weather.christmas_snow$snowfall), ]

# yearly plot for years it snowed

with_snow %>%
  ggplot(aes(x=YEAR, y=snowfall)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="steelblue", size=1.25) +
    geom_point(na.rm = TRUE, fill="#00edfa", shape = 21) +
    geom_label_repel(aes(label=YEAR), data=with_snow) +
    labs(title="Years With Snowfall on Christmas Eve/Christmas Day", x="Year", y="Snowfall (inches)")
```

It has snowed on `r nrow(with_snow)` out of `r 2021-1869` Christmases.  

The largest snowfall recorded on Christmas/Christmas Eve was in `r largest_snowfall$YEAR` when it snowed `r largest_snowfall$snowfall` inches!!  





# Exploratory Data Analysis

#### Importing Weather Data
```{r }

weather_data <- data.frame(read.csv("NYC_weather_1869_2022.csv"))

#head(weather_data, 5)

```

##### Install Packages

```{r}
#install.packages("ggplot2")
#install.packages("reshape2")
#install.packages("gridExtra")
#install.packages("ggpmisc")

```

#### Separating Date into Day/Mont/Year columns

```{r}

weather_data$DATE <- as.Date(weather_data$DATE)

weather_data$YEAR <- as.numeric(format(weather_data$DATE, "%Y"))
weather_data$MONTH<- as.factor(format(weather_data$DATE, "%m"))
weather_data$DAY <- format(weather_data$DATE, "%d")

#head(weather_data, 5)

```

I broke out the date column into separate Day, Month, and Year columns. I did this to make the data easier to manipulate on a month or year level.

#### Subsetting Data and Finding Monthly Averages for Temp, Precip, and Snowfall

``` {r}

weather_100yr <- subset(weather_data, weather_data$YEAR > 1899 & weather_data$YEAR < 2022)

month_tmax_avg <- aggregate(TMAX ~ MONTH + YEAR, weather_data, mean)
month_tmin_avg <- aggregate(TMIN ~ MONTH + YEAR, weather_data, mean)

month_precip_avg <- aggregate(PRCP ~ MONTH + YEAR, weather_data, mean)
colnames(month_precip_avg)[3] <- "PRCP_AVG"

month_precip_total <- aggregate(PRCP ~ MONTH + YEAR, weather_data, sum)
colnames(month_precip_total)[3] <- "PRCP_TOT"

month_snow_avg <- aggregate(SNOW ~ MONTH + YEAR, weather_data, mean)
colnames(month_snow_avg)[3] <- "SNOW_AVG"

month_snow_total <- aggregate(SNOW ~ MONTH + YEAR, weather_data, sum)
colnames(month_snow_total)[3] <- "SNOT_TOT"


monthly_weather <- data.frame(month_tmax_avg, 
                                             month_tmin_avg$TMIN,
                                             month_precip_avg$PRCP_AVG,
                                             month_precip_total$PRCP_TOT,
                                             month_snow_avg$SNOW_AVG,
                                             month_snow_total$SNOT_TOT)
colnames(monthly_weather)[4:8] <- c("TMIN", "PRCP_AVG", "PRCP_TOT", "SNOW_AVG", "SNOW_TOT")

monthly_weather_100yr <- subset(monthly_weather, monthly_weather$YEAR > 1899 & monthly_weather$YEAR < 2022)


```

Found the monthly average for TMAX, TMIN, PRCP, and SNOW for all months from 1900-2022. Additionally found the total monthly PRCP and SNOW for all months from 1900 - 2022. 

#### Doing the same for Yearly Data

``` {r}

year_tmax_avg <- aggregate(TMAX ~ YEAR, weather_data, mean)
year_tmin_avg <- aggregate(TMIN ~ YEAR, weather_data, mean)

year_precip_avg <- aggregate(PRCP ~ YEAR, weather_data, mean)
colnames(year_precip_avg)[2] <- "PRCP_AVG"

year_precip_total <- aggregate(PRCP ~ YEAR, weather_data, sum)
colnames(year_precip_total)[2] <- "PRCP_TOT"

year_snow_avg <- aggregate(SNOW ~ YEAR, weather_data, mean)
colnames(year_snow_avg)[2] <- "SNOW_AVG"

year_snow_total <- aggregate(SNOW ~ YEAR, weather_data, sum)
colnames(year_snow_total)[2] <- "SNOT_TOT"


yearly_weather <- data.frame(year_tmax_avg, 
                             year_tmin_avg$TMIN, 
                             year_precip_avg$PRCP_AVG, 
                             year_precip_total$PRCP_TOT, 
                             year_snow_avg$SNOW_AVG, 
                             year_snow_total$SNOT_TOT)
colnames(yearly_weather)[3:7] <- c("TMIN", "PRCP_AVG", "PRCP_TOT", "SNOW_AVG", "SNOW_TOT")


yearly_weather_100yr <- subset(yearly_weather, yearly_weather$YEAR > 1899 & yearly_weather$YEAR < 2022)

library(reshape2)

melt_year_data <- melt(yearly_weather_100yr, id.vars = "YEAR")

```

Found yearly averages for TMAX, TMIN, PRCP, and SNOW. Additionally found the yearly totals for PRCP and SNOW. 

### Plotting the Data to Visual Monthly & Yearly Trends

#### Yearly TMAX and TMIN
```{r}
library(ggplot2)

yearly_tmax_scatter <- ggplot(yearly_weather_100yr, aes(x = YEAR, y = TMAX)) + 
  geom_point(color = "coral2") +
  labs(x = "Year", y = "Temperature (F)", title = "Yearly Average Maximum Temp")

yearly_tmax_scatter

yearly_tmin_scatter <- ggplot(yearly_weather_100yr, aes(x = YEAR, y = TMIN)) + 
  geom_point(color = "cornflowerblue") +
  labs(x = "Year", y = "Temperature (F)", title = "Yearly Average Minimum Temp")

yearly_tmin_scatter

#Combined Tmax - Tmin plot
max_min_sub <- subset(melt_year_data, variable == "TMAX" | variable == "TMIN")

yearly_max_min_scatter <- ggplot(max_min_sub, aes(x = YEAR, y = value, col = variable)) + 
  geom_point() +
  labs(x = "Year", y = "Temperature (F)", title = "Yearly Average Maximum and Minimum Daily Temperatures in Central Park, NYC, from 1900-2022")

yearly_max_min_scatter

#tmax_scatter <- ggplot(monthly_weather, aes(x = YEAR, y = TMAX, col = MONTH)) +
#  geom_point() +
#  labs(x = "Year", y = "Maximum Daily Temperature (˚F)", title = "Seasonal Max Temperatures in Central Park, NYC, Over Time") +
#  scale_x_continuous(breaks = seq(1900, 2022, 10), minor_breaks = NULL)
  #guides(x = guide_axis(angle = 90, n.dodge = 5))
  
#tmax_scatter

```


#### Yearly PRCP and SNOW Plots
```{r}
yearly_prcp_scatter <- ggplot(yearly_weather_100yr, aes(x = YEAR, y = PRCP_TOT)) + 
  geom_point(color = "darkorchid") +
  labs(x = "Year", y = "Total Precipitation (in)", title = "Total Yearly Precipitation")      

yearly_prcp_scatter

yearly_snow_scatter <- ggplot(yearly_weather_100yr, aes(x = YEAR, y = SNOW_TOT)) + 
  geom_point(color = "azure4") +
  labs(x = "Year", y = "Total Snowfall (in)", title = "Total Yearly Snowfall")

yearly_snow_scatter

```

The above code generates 5 plots. One each for TMAX, TMIN, PRCP, and SNOW. Then an additional plot that plots both TMAX and TMIN on the same graph. 


```{r}

library(gridExtra)

grid.arrange(yearly_tmax_scatter, yearly_tmin_scatter,
             yearly_prcp_scatter, yearly_snow_scatter,
             nrow = 2, 
             top = "Weather Patterns in Central Park, NYC, from 1900-2021")

```

The quad chart above visualizes the four key weather patterns pulled from this data set. **Top left chart**: Max Daily Temps over time. There appears to be a clear trend of increasing max temperature over time.
**Top right chart**: Min daily temperatures over time. There appears to be a similar trend of increasing min temp over time, however the change is not as drastic as max temps.
**Bottom left chart**: Total precipitation per year. There does not appear to be any clear trend, however there appears to be more variation over the past 50 years.
**Bottom right chart**: Total snowfall per year. There does not appear to be any clear trend. Potentially interesting to see if any discernable difference in trend from yearly precipitation.


# Statistical Analysis

## Question 1:
Are there statistically measurable changes in weather patterns (e.g., temperature and precipitation levels) over time in New York City over this window?

- What statistical test?
- Does this require a linear model?



Using a Linear Regression to estimate the linear trend and statistical significance over time.

### TMAX
```{r}
#install.packages("gtsummary")

lm_tmax_yr <- lm(TMAX ~ YEAR, data = yearly_weather_100yr)
#lm_tmax_mtyr <- lm(TMAX ~ YEAR + MONTH, data = monthly_weather_100yr)
#lm_tmax_day <- lm(TMAX ~ YEAR, data = weather_100yr)

summary(lm_tmax_yr)
#summary(lm_tmax_mtyr)
#summary(lm_tmax_day)


coef_tmax_yr <- c(coefficients(lm_tmax_yr), summary(lm_tmax_yr)$adj.r.squared)
coef_tmax_yr


```

The table above shows the slope and intercept of the TMAX linear model based on two different approaches.`coef_tmax_day` uses the raw daily TMAX for each year in the dataset while `coef_tmax_yr` uses the yearly tmax averages. 

```{r}

library(ggpmisc)

yr_tmax_scatter_line <- yearly_tmax_scatter+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              color = "black",
              show.legend = F)

yr_tmax_scatter_line



```

The regression line added to the yearly TMAX scatterplot. The slope of the line is `r coef_tmax_yr[2]`. This implies the average daily maximum temperature is increasing by `r coef_tmax_yr[2]` degrees F per year, or by about `r coef_tmax_yr[2]*10` degrees F per decade.  


### TMIN

```{r}
lm_tmin_yr <- lm(TMIN ~ YEAR, data = yearly_weather_100yr)
summary(lm_tmin_yr)

#lm_tmin_day <- lm(TMIN ~ YEAR, data = weather_100yr)
#summary(lm_tmin_day)


#coef_tmin_yr <- coefficients(lm_tmin_yr)
#coef_tmin_day <- coefficients(lm_tmin_day)


```

The table above shows the slope and intercept of the TMIN linear model based on two different approaches.`coef_tmin_day` uses the raw daily TMIN for each year in the dataset while `coef_tmin_yr` uses the yearly TMIN averages. 

```{r}


yr_tmin_scatter_line <- yearly_tmin_scatter +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              color = "black",
              show.legend = F)

yr_tmin_scatter_line

```


### PRCP
```{r}

lm_prcp_yr <- lm(PRCP_TOT ~ YEAR, data = yearly_weather_100yr)
lm_prcp_month <- lm(PRCP_TOT ~ YEAR + MONTH, data = monthly_weather_100yr)
lm_prcp_avg <- lm(PRCP_AVG ~ YEAR + MONTH, data = monthly_weather_100yr)

lm_prcp_day <- lm(PRCP ~ YEAR, data = weather_100yr)
summary(lm_prcp_month)
summary(lm_prcp_yr)
summary(lm_prcp_avg)

coef_prcp_yr <- c(coefficients(lm_prcp_yr), summary(lm_prcp_yr)$r.square)
coef_prcp_yr
#coef_prcp_day <- coefficients(lm_tmax_day)
#coef_table_prcp <- cbind(coef_prcp_day, coef_prcp_yr)
#coef_table_prcp

```

The table above shows the slope and intercept of the TMAX linear model based on two different approaches.`

```{r}
yr_prcp_scatter_line <- yearly_prcp_scatter+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              color = "black",
              show.legend = F) + 
  labs(title = "Total Yearly Precipitation")


yr_prcp_scatter_line

```

### SNOW
```{r}

lm_snow_yr <- lm(SNOW_TOT ~ YEAR, data = yearly_weather_100yr)
summary(lm_snow_yr)

coef_snow_yr <- coefficients(lm_snow_yr)
coef_snow_yr

```

The table above shows the slope and intercept of the TMAX linear model based on two different approaches.`coef_tmax_day` uses the raw daily TMAX for each year in the dataset while `coef_tmax_yr` uses the yearly tmax averages. 

```{r}


yr_snow_scatter_line <- yearly_snow_scatter+
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              color = "black",
              show.legend = F)+
  labs(title = "Total Yearly Snowfall") 


yr_snow_scatter_line

```

```{r}
library(gridExtra)

grid.arrange(yr_prcp_scatter_line, yr_snow_scatter_line,
             yr_tmax_scatter_line, yr_tmin_scatter_line,
             nrow = 2, 
             top = "Weather Patterns in Central Park, NYC, from 1900-2021")



```



## Question 3:
Do the observed changes in the weather data from New York City align with the documented rise in global air temperatures over the last century?

- Compare rate of change observed in NYC to the global rate
- What stat test? 
- Earth’s temperature has risen by 0.14° Fahrenheit (0.08° Celsius) per decade since 1880, but the rate of warming since 1981 is more than twice that: 0.32° F (0.18° C) per decade. (per climate.gov)


To compare the slope of TMAX from the NYC data set to the known global temp change, will look at confidence interval of the coefficients from the linear model.

```{r}

weather1880 <- subset(monthly_weather, monthly_weather$YEAR > 1879 & monthly_weather$YEAR < 2022)
yearly_1880 <- subset(yearly_weather, yearly_weather$YEAR > 1879 & yearly_weather$YEAR < 2022)


yr1880_scatter <- ggplot(yearly_1880, aes(x = YEAR, y = TMAX)) + 
  geom_point() +
  labs(x = "Year", y = "Temperature (F)", title = "Yearly Average Maximum Temp")


yr1880_scatter

yr1880_line <- yr1880_scatter +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              col = "black") +
  labs(title = "NYC vs Global Temperature Change since 1880")
yr1880_line

yr1880_global <- yr1880_line + 
  geom_abline(aes(intercept = 33.2, slope = 0.014, col = "red")) +
  scale_color_discrete(name = "Locale", labels = c("Global"))


yr1880_global

lm1880 <- lm(TMAX ~ YEAR + MONTH, data = weather1880)
lm_yr1880 <- lm(TMAX ~ YEAR, data = yearly_1880)

summary(lm1880)
summary(lm_yr1880)


```


```{r}
weather1981 <- subset(monthly_weather, monthly_weather$YEAR > 1980 & monthly_weather$YEAR < 2022)
yearly_1981 <- subset(yearly_weather, yearly_weather$YEAR > 1980 & yearly_weather$YEAR < 2022)
all_1981 <- subset(weather_data, weather_data$YEAR > 1980 & weather_data$YEAR < 2022)

yr1981_scatter <- ggplot(yearly_1981, aes(x = YEAR, y = TMAX)) + 
  geom_point() +
  labs(x = "Year", y = "Temperature (F)", title = "Yearly Average Maximum Temp")
yr1981_scatter


yr1981_line <- yr1981_scatter +
  stat_smooth(method = "lm",
              formula = y ~ x,
              geom = "smooth",
              col = "black") +
  labs(title = "NYC vs Global Temperature Change since 1981")
yr1981_line

yr1981_global <- yr1981_line + 
  geom_abline(aes(intercept = -0.6, slope = 0.032, col = "red")) +
  scale_color_discrete(name = "Locale", labels = c("Global"))

yr1981_global

lm1981 <- lm(TMAX ~ YEAR + MONTH, data = weather1981)
lm_yearly1981 <- lm(TMAX ~ YEAR, data = yearly_1981)
lm_all1981 <- lm(TMAX ~ YEAR + MONTH, data = all_1981)
summary(lm1981)
summary(lm_yearly1981) 
summary(lm_all1981)

```


## Question 2:
Are there any notable (and statistically significant) changes in weather patterns after/during the pandemic lockdown (perhaps due to changes in commuting patterns)?

- Compare 10 years prior to 2020 to 2 years after?
- Chi-square test or 2-sample t-test?


```{r}
library(tibble)

covid_weather <- subset(monthly_weather, YEAR > 2010 & YEAR < 2021 &
                          (MONTH == "04" |MONTH == "05" | MONTH == "06" |MONTH == "07" |
                              MONTH == "08"))

covid_weather <- covid_weather %>%
  add_column(Lockdown = 
               ifelse(.$YEAR < 2020, "Pre-Lockdown", "Post-Lockdown"),
             .after = "YEAR")

covid_weather <- covid_weather %>%
  add_column(Season = 
               ifelse(.$MONTH == "04" | .$MONTH == "05", "Spring", "Summer"),
             .after = "YEAR")


covid_seasons <- aggregate(TMAX ~ Season + Lockdown, covid_weather, mean)
covid_seasons$Lockdown <- factor(covid_seasons$Lockdown, levels = c('Pre-Lockdown', 'Post-Lockdown'))

covid_bar2 <- ggplot(covid_seasons, aes(x = Season, y = TMAX, fill = Lockdown, decreasing = TRUE)) +
  geom_col(position = "dodge") +
  labs(title = "Seasonal Average Daily Rainfall Pre and Post COVID-19 Lockdown")

covid_bar2


```

```{r}

covid_spring_summer <- subset(weather_data,
                           YEAR > 2010 &
                           YEAR <= 2020 &
                             (MONTH == "04" |MONTH == "05" | MONTH == "06" |
                                MONTH == "07" | MONTH == "08"))
covid_spring_summer_ld <- covid_spring_summer %>%
  add_column(Lockdown = 
               ifelse(.$YEAR < 2020, "Pre-Lockdown", "Post-Lockdown"),
             .after = "YEAR")
covid_spring_summer_ld$Lockdown <- factor(covid_spring_summer_ld$Lockdown, levels = c('Pre-Lockdown', 'Post-Lockdown'))

covid_ld_season <- covid_spring_summer_ld %>%
  add_column(Season = 
               ifelse(.$MONTH == "04" | .$MONTH == "05", "Spring", "Summer"),
             .after = "YEAR")

covid_boxplot <- ggplot()+
  geom_boxplot(data = covid_ld_season, aes(y = TMAX, x = Season, fill = Lockdown)) +
  labs(title = "Temperature Ranges Pre- and Post- COVID Lockdown")

covid_boxplot



```


```{r}
pre_spring <- subset(weather_data,
                           YEAR > 2010 & YEAR < 2020 &
                           (MONTH == "04" |MONTH == "05"))


pre_summer <- subset(weather_data,
                           YEAR > 2010 & YEAR < 2020 &
                           (MONTH == "06" |MONTH == "07" |
                            MONTH == "08"))


covid_lockdown_spring <- subset(weather_data,
                           YEAR == 2020 &
                           (MONTH == "04" |MONTH == "05"))


covid_lockdown_summer <- subset(weather_data,
                           YEAR == 2020 &
                           (MONTH == "06" |MONTH == "07" |
                            MONTH == "08"))

library(reshape2)

cov_spring_ttest <- t.test(pre_spring$TMAX, covid_lockdown_spring$TMAX)
cov_spring_ttest

cov_summer_ttest <- t.test(pre_summer$TMAX, covid_lockdown_summer$TMAX)
cov_summer_ttest



```

##Looking at daily Temperature trends, Linear regressions, multiple linear regressions, and data from an additional NYC Location (JFK Airport)  

```{r}
library(ggplot2)
library(ezids)
library(dplyr)
```

##EDA with NYC Weather Data : import and cleaning

```{r, results="asis"}
#loading packages and data 

NYweath <- data.frame(read.csv("NYC_weather_1869_2022.csv"))
xkablesummary(NYweath)
#str(NYweath)                      
#cablehead(NYweath, n=5)
#tail(NYweath, n=2)
nrow(NYweath)

```

##More cleaning  

```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
NYweath$DATE <- as.Date(NYweath$DATE)
NYweath$day <- format(NYweath$DATE, format="%d")
NYweath$month <- format(NYweath$DATE, format="%m")
NYweath$year <- format(NYweath$DATE, format="%Y")

#converting temperature observations to numerics
NYweath$TMAX <- as.numeric(NYweath$TMAX)
NYweath$TMIN <- as.numeric(NYweath$TMIN)
NYweath$TAVG <- as.numeric(NYweath$TAVG)
NYweath$year <- as.numeric(NYweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data that are only date- and temperature-related
NYweath_sub <- subset(NYweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(NYweath_sub)
```

## First attempt to visualize temperature trends

Let's try to plot the max temp recorded each day in Central Park over time.

```{r}
NYweath_sub %>% 
  ggplot(aes(x = DATE, y = TMAX)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

And the minimum.

```{r}
NYweath_sub %>% 
  ggplot(aes(x = DATE, y = TMIN)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

What a mess! Wait, of course this will look like a mess, because temperature oscillates with the seasons! Let's make sure this oscillation is present in the data by looking at just a 2-year window of minimum daily temperatures from 2018 to the present.

```{r}
Recent_weath <- subset(NYweath_sub, year > 2017) 
Recent_weath %>% 
  ggplot(aes(x = DATE, y = TMIN)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

How reassuring! Now, let's look at the maximum daily temperatures.

```{r}
Recent_weath <- subset(NYweath_sub, year > 2017) 
Recent_weath %>% 
  ggplot(aes(x = DATE, y = TMAX)) +
    geom_point() +
    ggtitle("Maximum Daily Temperature in Central Park")
```

Cool. We could fit these data to sine curves to determine the amplitudes or center points of oscillation along with confidence intervals, and compare them to the results for way earlier in the 20th or 19th centuries, to see if there are statistical differences in the curve fit parameters across the 100-year window.

What about trends in daily temperature range (daily high minus daily low)?

```{r}
Recent_weath$rang <- (Recent_weath$TMAX - Recent_weath$TMIN)

Recent_weath %>% 
  ggplot(aes(x = DATE, y = rang)) +
  geom_point() + 
  labs(
    title = "Daily Temperature Range in Central Park",
    x = "Date",
    y = "Daily high minus low temp")
    
```

Interesting... I do see a trend here too: the greatest daily variation in temperature in Central Park appears to have some periodicity. We could compute the average daily temperature variation for each month, perhaps over the 100 year interval, to see if some months have statistically higher daily temperature variations.

## Linear regression of TMAX \~ year

```{r}
maxTfit0 <- lm(formula = TMAX ~ year, data = NYweath_sub )
summary(maxTfit0)
```

## Multiple linear regression with month as a categorical variable

Trying a fit on TMAX for all data using month as a factor variable.

```{r}
maxTfit1 <- lm(formula = TMAX ~ year + month, data = NYweath_sub )
summary(maxTfit1)

```

The multiple r-squared (and adjusted r-squared) fits are much better here, at 0.775. This may simply be because there are many more data points than when using monthly averages. So-- that full mess of data was all useful!

Here are the equations:

January: -23.3 + 0.0316 `year`\
February: -23.3+ 1.38 + 0.0316 `year`\
March: -23.3 + 9.55 + 0.0316 `year`\
April: -23.3 + 21.2 + 0.0316 `year`\
May: -23.3 + 32.3 + 0.0316 `year`\
June: -23.3 + 41.1 + 0.0316 `year`\
July: -23.3 + 46.1 + 0.0316 `year`\
August: -23.3 + 44.1 + 0.0316 `year`\
September: -23.3 + 37.4 + 0.0316 `year`\
October: -23.3 + 26.2 + 0.0316 `year`\
November: -23.3 + 14.1 + 0.0316 `year`\
December: -23.3 + 3.71 + 0.0316 `year`

On average, maximum daily temperatures in this location in central park have been increasing by 0.0316 degrees Fahrenheit per year over the time frame of the data set (1869 to 2022).

## Linear regression of TMAX \~ year \* month

```{r}
maxTfit2 <- lm(formula = TMAX ~ year*month, data = NYweath_sub )
summary(maxTfit2)

```

The equations, including interaction terms, are:

January: -2.998 + 0.01814 `year`\
February: -48.49 + -2.998 (0.02563 + 0.01814) `year`\
March: -60.35 + -2.998 + (0.03592 + 0.01814) `year`\
April: -37.07 + -2.998 + (0.02993 + 0.01814) `year`\
May: 19.85 + -2.998 + (0.00641 + 0.01814) `year`\
June: 47.84 + -2.998 + (0.02563 + 0.01814) `year`\
July: 39.37 + -2.998 + (0.0328 + 0.01814) `year`\
August: 22.69 + -2.998 + (0.01103 + 0.01814) `year`\
September: 26.84 + -2.998 + (0.00541 + 0.01814) `year`\
October: 16.26 + -2.998 + (0.00511 + 0.01814) `year`\
November: -31.07 + -2.998 + (0.02330 + 0.01814) `year`\
December: -36.39 + -2.998 + (0.02061 + 0.01814) `year`

Essentially the same r-squared here. The interaction terms are significant for February, March, April, August, November, and December, meaning that the change in temperature over time is likely different for these months. The warming is greatest in February, March, April, November, and December.

Now, let's subset to 1900 to 2022.

## Looking only at data from 1900 on

```{r}
NYweath_00 <- subset (NYweath_sub, year > 1899) # subsetting to 20th century and later
xkabledplyhead(NYweath_00)

```

## Linear regression of TMAX ~ year from 1900 on.

```{r}
maxTfit00_0 <- lm(formula = TMAX ~ year, data = NYweath_00 )
summary(maxTfit00_0)
```

## Linear regression of TMAX ~ year + month from 1900 on.

```{r}
maxTfit00_1 <- lm(formula = TMAX ~ year + month, data = NYweath_00 )
summary(maxTfit00_1)

```

## Linear regression of TMAX ~ year * month from 1900 on.

```{r}
maxTfit00_2 <- lm(formula = TMAX ~ year * month, data = NYweath_00 )
summary(maxTfit00_2)

```

## Plotting daily max temp since 2000 with color as month

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("01" = "purple4",
                                  "02" = "purple",
                                  "03" = "dodgerblue",
                                  "04" = "cyan4",
                                  "05" = "yellow",
                                  "06" = "darkgoldenrod1",
                                  "07" = "red",
                                  "08" = "orange", 
                                  "09" = "yellow",
                                  "10" = "green",
                                  "11" = "turquoise",
                                  "12" = "steelblue")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```
## Plotting the same data, but removing all but January and July

```{r}
ggplot(NYweath_00, aes(x = year, y = TMAX, color = month)) +
    geom_point() +
    scale_color_manual(values = c("01" = "purple4",
                                   "07" = "red"), na.value = NA) +
    geom_abline(aes(intercept = -11.05508, slope = 0.02539), col = "black", size = 1) + 
    geom_abline(aes(intercept = 34.98295, slope = 0.02539), col = "black", size = 1) +
  
    labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park")
#    stat_smooth(method = "lm",
#            formula = y ~ x,
#            geom = "smooth",
#            color = "black",
#            show.legend = F)
```
###Greyscale daily max temps from 1900-2022

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("01" = "gray35",
                                  "02" = "gray35",
                                  "03" = "gray35",
                                  "04" = "gray35",
                                  "05" = "gray35",
                                  "06" = "gray35",
                                  "07" = "gray35",
                                  "08" = "gray35", 
                                  "09" = "gray35",
                                  "10" = "gray35",
                                  "11" = "gray35",
                                  "12" = "gray35")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```


## Subsetting from 1948 on

```{r}
NYweath_48 <- subset (NYweath_sub, year > 1947) # subsetting to 1948 and later for comparison with JFK data
xkabledplyhead(NYweath_48)

```

## Linear regression of TMAX \~ year from 1949 on.

```{r}
maxTfit48_0 <- lm(formula = TMAX ~ year, data = NYweath_48 )
summary(maxTfit48_0)
```

## Linear regression of TMAX \~ year + month from 1948 on.

```{r}
maxTfit48_1 <- lm(formula = TMAX ~ year + month, data = NYweath_48 )
summary(maxTfit48_1)

```

## Linear regression of TMAX \~ year \* month from 1948 on.

```{r}
maxTfit48_2 <- lm(formula = TMAX ~ year*month, data = NYweath_48 )
summary(maxTfit48_2)

```



I wonder if we'll get something different if we look at weather in a more built environment. Starting a new RMD file for this (JFK airport).  

##JFK Data!

```{r}
JFKweath <- data.frame(read.csv("JFK_weather_1949-2022.csv"))
xkablesummary(JFKweath)
```
```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
JFKweath$DATE <- as.Date(JFKweath$DATE)
JFKweath$day <- format(JFKweath$DATE, format="%d")
JFKweath$month <- format(JFKweath$DATE, format="%m")
JFKweath$year <- format(JFKweath$DATE, format="%Y")

#converting temperature observations to numerics
JFKweath$TMAX <- as.numeric(JFKweath$TMAX)
JFKweath$TMIN <- as.numeric(JFKweath$TMIN)
JFKweath$TAVG <- as.numeric(JFKweath$TAVG)
JFKweath$year <- as.numeric(JFKweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data
JFKweath_sub <- subset(JFKweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(JFKweath_sub)
```

## Multiple linear regression for JFK from 1948 on  

```{r}
jmaxTfit1 <- lm(formula = TMAX ~ year + month, data = JFKweath_sub )
summary(jmaxTfit1)

```