---
title: "Project writeup"
author: "Emily"
date: "2022-10-29"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

output: html_document: code_folding: hide number_sections: false toc:
yes toc_depth: 3 toc_float: yes pdf_document: toc: yes toc_depth: '3'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

#loading packages 
library(ggplot2)
library(ezids)
library(dplyr)
```

## Emily's starter Code

```{r, results="asis"}
#loading data 

NYweath <- data.frame(read.csv("data/NYC_weather_1869_2022.csv"))
#xkablesummary(NYweath)
JFKweath <- data.frame(read.csv("data/JFK_weather_1949-2022.csv"))
#xkablesummary(JFKweath)

```

```{r, results="asis"}
#loading packages and data 

NYweath <- data.frame(read.csv("data/NYC_weather_1869_2022.csv"))
#xkablesummary(NYweath)
#str(NYweath)                      
#cablehead(NYweath, n=5)
#tail(NYweath, n=2)
#nrow(NYweath)

```

```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
NYweath$DATE <- as.Date(NYweath$DATE)
NYweath$day <- format(NYweath$DATE, format="%d")
NYweath$month <- format(NYweath$DATE, format="%m")
NYweath$year <- format(NYweath$DATE, format="%Y")

#converting temperature observations to numerics
NYweath$TMAX <- as.numeric(NYweath$TMAX)
NYweath$TMIN <- as.numeric(NYweath$TMIN)
NYweath$TAVG <- as.numeric(NYweath$TAVG)
NYweath$year <- as.numeric(NYweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data that are only date- and temperature-related
NYweath_sub <- subset(NYweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(NYweath_sub)
```

##Introduction

As scientists, our team is curious about how humans can affect the
natural environment. We wondered if weather conditions in a highly
populated region might correlate with notable human activities on a
global or local scale.

There is a variety of literature available on global climate change,
suggesting that the Earth's surface temperature is increasing, on
average over time, in a way that threatens many lifeforms on our planet.
This warming is primarily due to human activity: the combustion of
fossil fuels resulting in the emission of carbon dioxide into the
atmosphere. We decided to consider local weather trends in a specific
urban area to see what we might learn.

After some search, we discovered that historical daily weather records
from locations around the United States are made publicly availably from
the National Oceanic and Atmospheric Administrative via their website:
<https://www.ncdc.noaa.gov/cdo-web>. We found that a data station in the
middle of Central Park in New York City had made more than 56,000 daily
weather observations dating back to 1869. The variables observed
included daily maximum temperature in degrees Fahrenheit (TMAX), daily
minimum temperature in degrees Fahrenheit (TMIN), daily precipitation in
inches (PRCP), and daily snowfall (in inches). We decided to analyze
these data to see what trends we might uncover. Our data span the days
from February 28, 1869, to September 26, 2022.

Temperature data from Central Park have been studied in the past, and
warming trends were observed; we wanted to see what we might discover on
our own, from analysis of the original data.

## Our Research Questions

Given that Central Park is an oasis in the middle of a city environment,
we wondered whether any weather trends correlate with major human
activities-- both globally and locally. To explore these possibilities,
we formulated the following preliminary questions:

1.  Are there statistically measurable changes in weather patterns
    (e.g., temperature and precipitation levels) over time in New York
    City?\
2.  Do temperature trends in New York City align with the documented
    rise in average global temperatures over the last century?\
3.  Are there any notable (and statistically significant) changes in
    weather patterns after/during the pandemic lockdown (perhaps due to
    changes in commuting patterns)?

To dig in, we began with temperature.

## Temperature - stat tests of temp trends over time

## Chris's linear model - blurb

## Emily's linear model

```{r}
maxTfit0 <- lm(formula = TMAX ~ year, data = NYweath_sub )
summary(maxTfit0)
```

```{r}
maxTfit1 <- lm(formula = TMAX ~ year + month, data = NYweath_sub )
summary(maxTfit1)

xkablevif(maxTfit1)
```

```{r}
maxTfit2 <- lm(formula = TMAX ~ year*month, data = NYweath_sub )
summary(maxTfit2)
xkablevif(maxTfit2)

```

## Looking only at data from 1900 on

```{r}
NYweath_00 <- subset (NYweath_sub, year > 1899) # subsetting to 20th century and later
xkabledplyhead(NYweath_00)

```

## Linear regression of TMAX \~ year from 1900 on.

```{r}
maxTfit00_0 <- lm(formula = TMAX ~ year, data = NYweath_00 )
summary(maxTfit00_0)
```

## Linear regression of TMAX \~ year + month from 1900 on.

```{r}
maxTfit00_1 <- lm(formula = TMAX ~ year + month, data = NYweath_00 )
summary(maxTfit00_1)

```

## Linear regression of TMAX \~ year \* month from 1900 on.

```{r}
maxTfit00_2 <- lm(formula = TMAX ~ year * month, data = NYweath_00 )
summary(maxTfit00_2)

```

## Plotting daily max temp since 2000 with color as month

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("01" = "purple4",
                                  "02" = "purple",
                                  "03" = "dodgerblue",
                                  "04" = "cyan4",
                                  "05" = "yellow",
                                  "06" = "darkgoldenrod1",
                                  "07" = "red",
                                  "08" = "orange", 
                                  "09" = "yellow",
                                  "10" = "green",
                                  "11" = "turquoise",
                                  "12" = "steelblue")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```

```{r}
ggplot(NYweath_00, aes(x = year, y = TMAX, color = month)) +
    geom_point() +
    scale_color_manual(values = c("01" = "purple4",
                                   "07" = "red"), na.value = NA) +
    geom_abline(aes(intercept = -11.05508, slope = 0.02539), col = "black", size = 1) + 
    geom_abline(aes(intercept = 34.98295, slope = 0.02539), col = "black", size = 1) +
  
    labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park")
#    stat_smooth(method = "lm",
#            formula = y ~ x,
#            geom = "smooth",
#            color = "black",
#            show.legend = F)
```

```{r}
NYweath_48 <- subset (NYweath_sub, year > 1947) # subsetting to 1948 and later for comparison with JFK data
xkabledplyhead(NYweath_48)

```

## Linear regression of TMAX \~ year from 1949 on.

```{r}
maxTfit48_0 <- lm(formula = TMAX ~ year, data = NYweath_48 )
summary(maxTfit48_0)
```

## Linear regression of TMAX \~ year + month from 1948 on.

```{r}
maxTfit48_1 <- lm(formula = TMAX ~ year + month, data = NYweath_48 )
summary(maxTfit48_1)

```

## Multiple linear regression for JFK from 1948 on

```{r}
jmaxTfit1 <- lm(formula = TMAX ~ year + month, data = JFKweath_sub )
summary(jmaxTfit1)

```

```{r}
NYweath_00 <- subset(NYweath_sub, year > 1899) 
# subsetting to 20th century and later
xkabledplyhead(NYweath_00)

```

```{r, results = "asis"}

maxTfit00_0 <- lm(formula = TMAX ~ year, data = NYweath_00 )  
summary(maxTfit00_0)  

maxTfit00_1 <- lm(formula = TMAX ~ year + month, data = NYweath_00 )

summary(maxTfit00_1)

maxTfit00_2 <- lm(formula = TMAX ~ year * month, data = NYweath_00 )

summary(maxTfit00_2)
```

```{r}

#Plotting all TMAX data from 1900 on
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("01" = "purple4",
                                  "02" = "purple",
                                  "03" = "dodgerblue",
                                  "04" = "cyan4",
                                  "05" = "yellow",
                                  "06" = "darkgoldenrod1",
                                  "07" = "red",
                                  "08" = "orange", 
                                  "09" = "yellow",
                                  "10" = "green",
                                  "11" = "turquoise",
                                  "12" = "steelblue")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```

```{r}
ggplot(NYweath_00, aes(x = year, y = TMAX, color = month)) +
    geom_point() +
    scale_color_manual(values = c("01" = "purple4",
                                   "07" = "red"), na.value = NA) +
    geom_abline(aes(intercept = -11.05508, slope = 0.02539), col = "black", size = 1) + 
    geom_abline(aes(intercept = 34.98295, slope = 0.02539), col = "black", size = 1) +
  
    labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park")
```

## Central Park v. Global average - Chris

## JFK data, comparable? - Emily

```{r}

#Subsetting central park data to observations from 1948 and later for comparison with JFK data

maxTfit48_0 <- lm(formula = TMAX ~ year, data = NYweath_48 )
summary(maxTfit48_0)

maxTfit48_1 <- lm(formula = TMAX ~ year + month, data = NYweath_48 )
summary(maxTfit48_1)

maxTfit48_2 <- lm(formula = TMAX ~ year*month, data = NYweath_48 )
summary(maxTfit48_2)

# Multiple linear regression for JFK (from 1948 on  
jmaxTfit1 <- lm(formula = TMAX ~ year + month, data = JFKweath_sub )
summary(jmaxTfit1)

```

How did thinking change? reframing of questioning?\
Additional data to test our hypothesis of causality

## Precipitation trends -Tejas' outlier analysis + snowfall at Christmas

## Linear models of precipitation trends over time - Chris

## Correlation between temp / precip -- Emily ... think about subsequent analyses

## Lockdown - Chris

## Conclusion- terse summary of findings, restate new questions from all sections, ideas for future analyses if you give us more money.

If we have time, summarize what we learned from each variable, and new
questions?
