---
title: "Project 1 EDA: NYC Weather Data"
author: "Emily"
date: "2022-10-01"
# date: "today"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 

library(ggplot2)
library(ezids)
library(dplyr)
```

##EDA with NYC Weather Data

```{r, results="asis"}
#loading packages and data 

NYweath <- data.frame(read.csv("NYC_weather_1869_2022.csv"))
xkablesummary(NYweath)
#str(NYweath)                      
#cablehead(NYweath, n=5)
#tail(NYweath, n=2)
nrow(NYweath)

```

```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
NYweath$DATE <- as.Date(NYweath$DATE)
NYweath$day <- format(NYweath$DATE, format="%d")
NYweath$month <- format(NYweath$DATE, format="%m")
NYweath$year <- format(NYweath$DATE, format="%Y")

#converting temperature observations to numerics
NYweath$TMAX <- as.numeric(NYweath$TMAX)
NYweath$TMIN <- as.numeric(NYweath$TMIN)
NYweath$TAVG <- as.numeric(NYweath$TAVG)
NYweath$year <- as.numeric(NYweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data that are only date- and temperature-related
NYweath_sub <- subset(NYweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(NYweath_sub)
```

## First attempt to visualize temperature trends

Let's try to plot the max temp recorded each day in Central Park over time.

```{r}
NYweath_sub %>% 
  ggplot(aes(x = DATE, y = TMAX)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

And the minimum.

```{r}
NYweath_sub %>% 
  ggplot(aes(x = DATE, y = TMIN)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

What a mess! Wait, of course this will look like a mess, because temperature oscillates with the seasons! Let's make sure this oscillation is present in the data by looking at just a 2-year window of minimum daily temperatures from 2018 to the present.

```{r}
Recent_weath <- subset(NYweath_sub, year > 2017) 
Recent_weath %>% 
  ggplot(aes(x = DATE, y = TMIN)) +
    geom_point() +
    ggtitle("Minimum Daily Temperature in Central Park")
```

How reassuring! Now, let's look at the maximum daily temperatures.

```{r}
Recent_weath <- subset(NYweath_sub, year > 2017) 
Recent_weath %>% 
  ggplot(aes(x = DATE, y = TMAX)) +
    geom_point() +
    ggtitle("Maximum Daily Temperature in Central Park")
```

Cool. We could fit these data to sine curves to determine the amplitudes or center points of oscillation along with confidence intervals, and compare them to the results for way earlier in the 20th or 19th centuries, to see if there are statistical differences in the curve fit parameters across the 100-year window.

What about trends in daily temperature range (daily high minus daily low)?

```{r}
Recent_weath$rang <- (Recent_weath$TMAX - Recent_weath$TMIN)

Recent_weath %>% 
  ggplot(aes(x = DATE, y = rang)) +
  geom_point() + 
  labs(
    title = "Daily Temperature Range in Central Park",
    x = "Date",
    y = "Daily high minus low temp")
    
```

Interesting... I do see a trend here too: the greatest daily variation in temperature in Central Park appears to occur in later winter/early spring. We could compute the average daily temperature variation for each month, perhaps over the 100 year interval, to see if some months have statistically higher daily temperature variations.

## Linear regression of TMAX \~ year

```{r}
maxTfit0 <- lm(formula = TMAX ~ year, data = NYweath_sub )
summary(maxTfit0)
```

## Multiple linear regression with month as a categorical variable

Trying a fit on TMAX for all data using month as a factor variable.

```{r}
maxTfit1 <- lm(formula = TMAX ~ year + month, data = NYweath_sub )
summary(maxTfit1)

```

The multiple r-squared (and adjusted r-squared) fits are much better here, at 0.775. This may simply be because there are many more data points than when using monthly averages. So-- that full mess of data was all useful!

Here are the equations:

January: -23.3 + 0.0316 `year`\
February: -23.3+ 1.38 + 0.0316 `year`\
March: -23.3 + 9.55 + 0.0316 `year`\
April: -23.3 + 21.2 + 0.0316 `year`\
May: -23.3 + 32.3 + 0.0316 `year`\
June: -23.3 + 41.1 + 0.0316 `year`\
July: -23.3 + 46.1 + 0.0316 `year`\
August: -23.3 + 44.1 + 0.0316 `year`\
September: -23.3 + 37.4 + 0.0316 `year`\
October: -23.3 + 26.2 + 0.0316 `year`\
November: -23.3 + 14.1 + 0.0316 `year`\
December: -23.3 + 3.71 + 0.0316 `year`

On average, maximum daily temperatures in this location in central park have been increasing by 0.0316 degrees Fahrenheit per year over the time frame of the data set (1869 to 2022).

## Linear regression of TMAX \~ year \* month

```{r}
maxTfit2 <- lm(formula = TMAX ~ year*month, data = NYweath_sub )
summary(maxTfit2)

```

The equations, including interaction terms, are:

January: -2.998 + 0.01814 `year`\
February: -48.49 + -2.998 (0.02563 + 0.01814) `year`\
March: -60.35 + -2.998 + (0.03592 + 0.01814) `year`\
April: -37.07 + -2.998 + (0.02993 + 0.01814) `year`\
May: 19.85 + -2.998 + (0.00641 + 0.01814) `year`\
June: 47.84 + -2.998 + (0.02563 + 0.01814) `year`\
July: 39.37 + -2.998 + (0.0328 + 0.01814) `year`\
August: 22.69 + -2.998 + (0.01103 + 0.01814) `year`\
September: 26.84 + -2.998 + (0.00541 + 0.01814) `year`\
October: 16.26 + -2.998 + (0.00511 + 0.01814) `year`\
November: -31.07 + -2.998 + (0.02330 + 0.01814) `year`\
December: -36.39 + -2.998 + (0.02061 + 0.01814) `year`

Essentially the same r-squared here. The interaction terms are significant for February, March, April, August, November, and December, meaning that the change in temperature over time is likely different for these months. The warming is greatest in February, March, April, November, and December.

Now, let's subset to 1900 to 2022.

## Looking only at data from 1900 on

```{r}
NYweath_00 <- subset (NYweath_sub, year > 1899) # subsetting to 20th century and later
xkabledplyhead(NYweath_00)

```

## Linear regression of TMAX ~ year from 1900 on.

```{r}
maxTfit00_0 <- lm(formula = TMAX ~ year, data = NYweath_00 )
summary(maxTfit00_0)
```

## Linear regression of TMAX ~ year + month from 1900 on.

```{r}
maxTfit00_1 <- lm(formula = TMAX ~ year + month, data = NYweath_00 )
summary(maxTfit00_1)

```

## Linear regression of TMAX ~ year * month from 1900 on.

```{r}
maxTfit00_2 <- lm(formula = TMAX ~ year * month, data = NYweath_00 )
summary(maxTfit00_2)

```

## Plotting daily max temp since 2000 colored by month

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.6) +
    scale_color_manual(values = c("01" = "purple4",
                                  "02" = "purple",
                                  "03" = "dodgerblue",
                                  "04" = "cyan4",
                                  "05" = "yellow",
                                  "06" = "darkgoldenrod1",
                                  "07" = "red",
                                  "08" = "orange", 
                                  "09" = "yellow",
                                  "10" = "green",
                                  "11" = "turquoise",
                                  "12" = "steelblue")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```
## Plotting the same data, but removing all but January and July

```{r}
ggplot(NYweath_00, aes(x = year, y = TMAX, color = month)) +
    geom_point() +
    scale_color_manual(values = c("01" = "purple4",
                                   "07" = "red"), na.value = NA) +
    geom_abline(aes(intercept = -11.05508, slope = 0.02539), col = "black", size = 1) + 
    geom_abline(aes(intercept = 34.98295, slope = 0.02539), col = "black", size = 1) +
  
    labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park")
#    stat_smooth(method = "lm",
#            formula = y ~ x,
#            geom = "smooth",
#            color = "black",
#            show.legend = F)
```
###Greyscale daily max temps from 1900-2022

```{r}
ggplot(NYweath_00, aes(x = DATE, y = TMAX, color = month)) + 
    geom_point(alpha = 0.7) +
    scale_color_manual(values = c("01" = "gray35",
                                  "02" = "gray35",
                                  "03" = "gray35",
                                  "04" = "gray35",
                                  "05" = "gray35",
                                  "06" = "gray35",
                                  "07" = "gray35",
                                  "08" = "gray35", 
                                  "09" = "gray35",
                                  "10" = "gray35",
                                  "11" = "gray35",
                                  "12" = "gray35")) +
      labs(
        x = "Year",
        y = "Maximum Daily Temperature",
        title = "Maximum Daily Temperature in Central Park") +
    xlab(label = "Year") +
    ylab(label = "Maximum daily temperature") +
    ggtitle(label = "Maximum Daily Temperature in Central Park") +
    stat_smooth(method = "lm",
            formula = y ~ x,
            geom = "smooth",
            color = "black",
            show.legend = F)
```


## Subsetting from 1948 on

```{r}
NYweath_48 <- subset (NYweath_sub, year > 1947) # subsetting to 1948 and later for comparison with JFK data
xkabledplyhead(NYweath_48)

```

## Linear regression of TMAX \~ year from 1949 on.

```{r}
maxTfit48_0 <- lm(formula = TMAX ~ year, data = NYweath_48 )
summary(maxTfit48_0)
```

## Linear regression of TMAX \~ year + month from 1948 on.

```{r}
maxTfit48_1 <- lm(formula = TMAX ~ year + month, data = NYweath_48 )
summary(maxTfit48_1)

```

## Linear regression of TMAX \~ year \* month from 1948 on.

```{r}
maxTfit48_2 <- lm(formula = TMAX ~ year*month, data = NYweath_48 )
summary(maxTfit48_2)

```



I wonder if we'll get something different if we look at weather in a more built environment. Starting a new RMD file for this (JFK airport).

```{r}
JFKweath <- data.frame(read.csv("JFK_weather_1949-2022.csv"))
xkablesummary(JFKweath)
```
```{r, r = "as is"}
#converting to R date format and adding columns for day, month, and year
JFKweath$DATE <- as.Date(JFKweath$DATE)
JFKweath$day <- format(JFKweath$DATE, format="%d")
JFKweath$month <- format(JFKweath$DATE, format="%m")
JFKweath$year <- format(JFKweath$DATE, format="%Y")

#converting temperature observations to numerics
JFKweath$TMAX <- as.numeric(JFKweath$TMAX)
JFKweath$TMIN <- as.numeric(JFKweath$TMIN)
JFKweath$TAVG <- as.numeric(JFKweath$TAVG)
JFKweath$year <- as.numeric(JFKweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

#creating a subset of the weather data
JFKweath_sub <- subset(JFKweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 
xkablesummary(JFKweath_sub)
```

## Multiple linear regression for JFK from 1948 on

```{r}
jmaxTfit1 <- lm(formula = TMAX ~ year + month, data = JFKweath_sub )
summary(jmaxTfit1)

```