---
title: "NYC Weather Data Preliminary EDA"
author: "Tejas Rawal"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# tidyverse includes ggplot2
library(tidyverse)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

# NYC Weather Data: Temperature analysis

```{r load data}
nyc_weather <- tibble(read.csv("NYC_weather_1869_2022.csv"))
# view structure
str(nyc_weather)
# get summary
xkablesummary(nyc_weather)
```

Convert DATE column from `char` to `Date` type
```{r}
nyc_weather <- nyc_weather %>%
  mutate(DATE = as.Date(DATE, format="%Y-%m-%d"))

# review dataset structure
xkablesummary(nyc_weather)
```

Add a MONTH column to the dataset. The month is extracted from the DATE.
```{r}
nyc_weather <- nyc_weather %>%
  mutate(MONTH = month(DATE, label=T))

# review dataset structure
xkablesummary(nyc_weather)
```


### Average daily temperature max and min by month, all time
```{r temperature analysis}
# group data by month
# and calculate average
nyc_weather_by_month <- nyc_weather %>%
  group_by(MONTH) %>%
  summarize(avg_temp_max = mean(TMAX, na.rm=T), avg_temp_min = mean(TMIN, na.rm=T), na.rm = T)

# Create line plot
# TODO: Add legend for each line
nyc_weather_by_month %>%
  ggplot(aes(x=MONTH, group=1)) +
  geom_line(aes(y=avg_temp_max), color="red") +
  geom_line(aes(y=avg_temp_min), color="steelblue", linetype=4) +
  labs(x="Month", y="Mean Temperate (°F)")
```

### Average daily temperature max and min by month, pre 1879-1900
```{r temperature analysis}
# filter weather data for years 1879-1900
nyc_weather %>%
  filter()

```

### Average daily temperature max and min by month, post 1900-1930
```{r temperature analysis}
```

### Average daily temperature max and min by month, post 1950-1980 century
```{r temperature analysis}
```

### Average daily temperature max and min by month, post 1950 century
```{r temperature analysis}
```

### Temperature vs Rain amount 2D Density plot
