---
title: "NYC Weather Data Preliminary EDA"
author: "Tejas Rawal"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
# tidyverse includes ggplot2
library(tidyverse)
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```

# NYC Weather Data: Temperature analysis

```{r load data}
nyc_weather <- tibble(read.csv("../data/NYC_weather_1869_2022.csv")) %>%
  dplyr::select(TMAX, TMIN, DATE, PRCP, SNOW)
# view structure
str(nyc_weather)
# get summary
xkablesummary(nyc_weather)
```

Convert DATE column from `char` to `Date` type
```{r}
nyc_weather <- nyc_weather %>%
  mutate(DATE = as.Date(DATE, format="%Y-%m-%d"))

# review dataset structure
xkablesummary(nyc_weather)
```

Add a MONTH column to the dataset. The month is extracted from the DATE.  
Subset data for DATE > 1900
```{r}
nyc_weather <- nyc_weather %>%
  filter(lubridate::year(DATE) >= 1900) %>%
  mutate(MONTH = lubridate::month(DATE, label=T))

str(nyc_weather)
# review dataset structure
xkablesummary(nyc_weather)
```

### Temperature Plots
```{r}

mean_tmax <- mean(nyc_weather$TMAX)
mean_tmin <- mean(nyc_weather$TMIN)

ggplot(nyc_weather) +
  geom_histogram(aes(x=TMAX, fill="Max Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmax, color="red", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmax + 9, y=2000, label=paste(round(mean_tmax, 2), "°F" ), angle=0, size=4, color="darkred") +
  geom_histogram(aes(x=TMIN, fill="Min Temp."), na.rm=TRUE, alpha=0.5, color="black", bins=100, binwidth=2) +
  geom_vline(xintercept=mean_tmin, color="blue", size=1, linetype=5, show.legend=FALSE) +
  annotate("text", x=mean_tmin - 9, y=2000, label=paste(round(mean_tmin, 2), "°F" ), angle=0, size=4, color="darkblue") +
  labs(title="Distribution of Minimum and Maximum Daily Temperatures", x="Temperature (°F)", y="Count") +
  scale_y_continuous(position = "right") +
  scale_fill_manual(name="Column", values=c("Max Temp."="red","Min Temp."="blue"))

nyc_weather %>%
  ggplot(aes(y=TMAX)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black", fill="red") +
  labs(title="TMAX Boxplot", y="Temperature (°F)")

nyc_weather %>%
  ggplot(aes(y=TMIN)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black", fill="steelblue") +
  labs(title="TMIN Boxplot", x="", y="Temperature (°F)")


qqnorm(nyc_weather$TMAX,
       main="QQ Plot TMAX",
       xlab="Theoretical Quantile",
       ylab="Temperature  (°F)",
       col="red")
qqline(nyc_weather$TMAX, lwd=1, lty=2)

qqnorm(nyc_weather$TMIN,
       main="QQ Plot TMIN",
       xlab="Theoretical Quantile",
       ylab="Temperature  (°F)",
       col="blue")
qqline(nyc_weather$TMIN, lwd=1, lty=2)
```

### Average daily temperature max and min by month, all time
```{r average daily temperature analysis}
# group data by month
# and calculate average
nyc_weather_by_month <- nyc_weather %>%
  group_by(MONTH) %>%
  summarize("Avg Max Temp." = mean(TMAX, na.rm=T),
            "Avg Min Temp." = mean(TMIN, na.rm=T),
            avg_prcp = mean(PRCP, na.rm=T),
            avg_snow = mean(SNOW, na.rm=T))

# side-by-side bar plot
nyc_weather_by_month %>%
  dplyr::select(MONTH, "Avg Max Temp.", "Avg Min Temp.") %>%
  gather(key="Value", value="Temp.", "Avg Max Temp.", "Avg Min Temp.") %>%
  ggplot(aes(x=MONTH, y=Temp., fill=Value)) +
  geom_col(na.rm=TRUE, alpha=0.5, color="black", position="dodge") +
  labs(title="Average Daily Temperature By Month", x="Month", y="Average Temperature (°F)") +
  scale_fill_manual(values=c("blue", "red"))
```


### Average percipitation and snow fall by month, all time
```{r}
nyc_weather_by_month %>%
  dplyr::select(MONTH, avg_prcp, avg_snow) %>%
  gather(key = "data_point", value="value", avg_prcp, avg_snow) %>%
  ggplot(aes(x=MONTH, y=value, fill=data_point)) +
    geom_col(na.rm=TRUE, alpha=0.7, color="black", position="dodge") +
    labs(x="Month", y="Amount (inches)")
```

## Categorical Analysis

For this analysis, I sought to subset the data into 30-50 year periods as such:  
* 1900-1940: Industrial Revolution  
* 1940-1980: Cold War Era  
* 1980-present: Modern Era

### 1900-1940
#### Temperature and Precipitation Boxplots
```{r}
# filter weather data for years 1900-1940
industrial_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1900, lubridate::year(DATE) < 1940)

str(industrial_era)

industrial_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1900-1940", x="Temperature (°F)", y="Count")

industrial_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1900-1940", x="Data Point", y="Temperature (°F)")

industrial_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1900-1940", x="Data Point", y="Amount (inches)")

```

### 1940-1980
#### Temperature and Precipitation Boxplots
```{r era-based temperature analysis}
coldwar_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1940, lubridate::year(DATE) < 1980)

str(coldwar_era)

coldwar_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1940-1980", x="Temperature (°F)", y="Count")

coldwar_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1940-1980", x="Data Point", y="Temperature (°F)")

coldwar_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1940-1980", x="Data Point", y="Amount (inches)")

```

### 1980 - present
#### Temperature and Precipitation Boxplots
```{r}
modern_era <- nyc_weather %>%
  dplyr::select(-MONTH) %>%
  filter(lubridate::year(DATE) >= 1980)

str(modern_era)

modern_era %>%
  ggplot() +
  geom_histogram(aes(x=TMAX, fill="TMAX"), na.rm=TRUE, alpha=0.8, fill="steelblue", bins=100, binwidth=2) +
  labs(title="Distribution of TMAX for Years 1980-Present", x="Temperature (°F)", y="Count")

modern_era %>%
  dplyr::select(TMAX, TMIN, DATE) %>%
  gather(key="data_point", value="value", TMAX, TMIN) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of TMAX and TMIN for Years 1980-Present", x="Data Point", y="Temperature (°F)")

modern_era %>%
  dplyr::select(PRCP, SNOW, DATE) %>%
  gather(key="data_point", value="value", PRCP, SNOW) %>%
  filter(value > 0.0) %>%
  ggplot(aes(y=value, fill=data_point)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  labs(title="Distribution of PRCP and SNOW for Years 1980-Present", x="Data Point", y="Amount (inches)")
```

### ANOVA test

#### Mean TMAX by era
```{r}
industrial_avg_tmax <- mean(industrial_era$TMAX)
coldwar_avg_tmax <- mean(coldwar_era$TMAX)
modern_avg_tmax <- mean(modern_era$TMAX)

industrial_avg_tmax
coldwar_avg_tmax
modern_avg_tmax
```

We can see above that there is an increase in average daily temperature max over the three eras.
$H_0:$ The average `TMAX` across the three eras is equal.  
$H_1:$ The average `TMAX` across the three eras is different.  
$\alpha$: `0.05`

```{r}
# modify data set by adding ERA column
industrial_era <- industrial_era %>%
  mutate(Era = 1)
coldwar_era <- coldwar_era %>%
  mutate(Era = 2)
modern_era <- modern_era %>%
  mutate(Era = 3)
# combine data frames
all_eras <- rbind(industrial_era, coldwar_era, modern_era)
# convert ERA column to factor/categorical
all_eras$Era <- factor(all_eras$Era, labels = c("Indst.", "Cold War", "Modern"))

# box plot
means <- aggregate(TMAX ~ Era, all_eras, mean)

all_eras %>%
  ggplot(aes(x=Era, y=TMAX, fill=Era)) +
  geom_boxplot(na.rm=TRUE, alpha=0.7, color="black") +
  stat_summary_bin(fun="mean", geom="point", shape=20, size=5, show.legend=FALSE) +
  geom_text(data = means, aes(label=round(TMAX, 2) , y=TMAX - 5)) +
  labs(title="Distribution of Daily Maximum Temperature By Era", x="Era", y="Temperature (°F)") +
  scale_fill_brewer(palette="Dark2")
```

```{r}
# run ANOVA
anova_res <- aov(TMAX ~ Era, data = all_eras)
summary(anova_res)
xkabledply(anova_res)
```

With a p-value of `r pf(51, 2, 44826, lower.tail=F)`, we can safely assume that the average daily maximum temperature between the 3 eras is different.

#### Tukey HSD

This tests will analyze the difference between means of the 3 eras:  

```{r}
tukeyAoV <- TukeyHSD(anova_res, conf.level=0.95)
tukeyAoV
plot(tukeyAoV, col="red", cex.axis=0.75)
```

### Outlier Check on Precipitation
```{r}
library(ggrepel)
# extract PRCP outlier rows into df variable
# remove zeros
# group variable by decade, getting sum for each decade
prcp_no_zeros <- nyc_weather %>%
  filter(PRCP > 0.0 & !is.na(PRCP))
prcp_boxplot <- boxplot(prcp_no_zeros$PRCP, ylab = "Temperature (°F)")

prcp_outliers <- boxplot.stats(prcp_no_zeros$PRCP)$out
prcp_outliers.index <- which(prcp_no_zeros$PRCP %in% c(prcp_outliers))
prcp_outliers_df <- prcp_no_zeros[prcp_outliers.index, ]

prcp_outliers_df.yearly <- prcp_outliers_df %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarize(prcp_total = sum(PRCP, na.rm=T), count = n())

# show top 5 years
top_5_years.prcp_total <- prcp_outliers_df.yearly %>%
  slice_max(order_by = prcp_total, n = 5)

prcp_outliers_df.yearly %>%
  ggplot(aes(x=YEAR, y=prcp_total)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="darkgreen") +
    geom_point(na.rm = TRUE, fill="#69b3a2", shape = 21) +
    geom_text_repel(aes(label=YEAR), data=top_5_years.prcp_total) +
    labs(x="Year", y="Total Yearly Amount (inches)") +
    scale_fill_brewer(palette="Spectral")

top_5_years.count <- prcp_outliers_df.yearly %>%
  slice_max(order_by = count, n = 5)

prcp_outliers_df.yearly %>%
  ggplot(aes(x=YEAR, y=count)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="darkgreen") +
    geom_point(na.rm = TRUE, fill="#69b3a2", shape = 21) +
    geom_text_repel(aes(label=YEAR), data=top_5_years.count) +
    labs(title="Precipitation Outlier Days By Year", x="Year", y="Days with Outliers")

# group into decades and create bar plot
prcp_outliers_df.decades <- prcp_outliers_df %>%
  mutate(decade = paste(floor(lubridate::year(DATE) / 10) * 10, "s", sep="")) %>%
  group_by(decade) %>%
  summarise(prcp_total = sum(PRCP, na.rm=T), count = n())

prcp_outliers_df.decades %>%
  ggplot(aes(group=1)) +
  geom_col(aes(x=decade, y=prcp_total), na.rm=TRUE, alpha=0.8, color="black", fill="darkgreen", position="identity") +
  geom_line(aes(x=decade, y=count), stat="identity", color="#69b3a2", size=1) +
  labs(title="Precipitation Outlier Days And Total Precipitation By Decade", x="Decade", y="Total Amount (inches)") +
  scale_y_continuous(sec.axis=sec_axis(~., name="Days with Outliers"))

prcp_outliers_df.decades %>%
  ggplot(aes(x=decade, y=count)) +
    geom_col(na.rm=TRUE, alpha=0.7, color="black", fill="#56B4E9", position="identity") +
    labs(x="Decade", y="Days with Outliers")
```

### Christmas Snow
```{r}
# sum snowfall from 24-25th Dec
nyc_weather.christmas_snow <- nyc_weather %>%
  filter(lubridate::month(DATE) == 12, (lubridate::day(DATE) == 24 | lubridate::day(DATE) == 25)) %>%
  mutate(YEAR = lubridate::year(DATE)) %>%
  group_by(YEAR) %>%
  summarise(snowfall = sum(SNOW, na.rm = T))

# how many years has there been snow?
with_snow <- nyc_weather.christmas_snow %>%
  filter(snowfall > 0.0)

# highest snowfall year?
largest_snowfall <- nyc_weather.christmas_snow[which.max(nyc_weather.christmas_snow$snowfall), ]

# yearly plot for years it snowed
top_5_years.snowfall <- with_snow %>%
  slice_max(order_by = snowfall, n = 5)

with_snow %>%
  ggplot(aes(x=YEAR, y=snowfall)) +
    geom_line(na.rm=TRUE, alpha=0.7, color="steelblue", size=1.25) +
    geom_point(na.rm = TRUE, fill="#00edfa", shape = 21) +
    geom_label_repel(aes(label=YEAR), data=with_snow) +
    labs(title="Years With Snowfall on Christmas Eve/Christmas Day", x="Year", y="Snowfall (inches)")
```

It has snowed on `r nrow(with_snow)` out of `r 2021-1869` Christmases.  

The largest snowfall recorded on Christmas/Christmas Eve was in `r largest_snowfall$YEAR` when it snowed `r largest_snowfall$snowfall` inches!!  